---
title: "SLE777 Assessment 4: Reproducible Differential Expression Report"
author: "Ivy Craw, Murray Pung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params: 
  group_id: "24"
  # Six assigned count files exactly as they appear in our allocation row
  condition_a_files: ["sample_024_counts.txt","sample_026_counts.txt","sample_029_counts.txt"]
  condition_b_files: ["sample_007_counts.txt","sample_017_counts.txt","sample_032_counts.txt"]
  counts_dir: "/home/shared/SLE777/A4_counts"
  gene_of_interest: "YDL022W"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 6)
set.seed(777)
# Load the packages you need. Record versions with sessionInfo() at the end.
library(clusterProfiler)
library(DESeq2)
library(ggplot2)
# What's magrittr used for?
library(magrittr)
library(apeglm)
library(pheatmap)
library(org.Sc.sgd.db)
library(AnnotationDbi)

# Suggested library green lit by unit chair for dynamic labelling
# OPTIONAL
library(ggrepel)

# Create figures directory if it does not already exist
dir.create("figures", showWarnings = FALSE)
```

This report analyses the cohort assigned to group ``r params$group_id``,
gene of interest ``r params$gene_of_interest``.

# 1. Count matrix and metadata

```{r matrix}
####################################
# Build the count matrix
####################################

# Combine six HTSeq-count files into one gene six sample matrix
# Samples: condition A versus condition B, for assessment purposes

# To compare gene expression between conditions, we need all six samples side by side in one table
# The end product of this is a count matrix
# One row per gene and one column per sample and a read count in every cell
# We also build a sample metadata table
# These two objects are required for differential expression analysis (DESeq2)

### Descriptions of samples
# Counts provided for assessment purposes
# No additional information provided

####################################
# The sample metadata table
####################################

# Create sample metadata table with sample identifiers and condition
sample_metadata <- data.frame(
  sample_id = c("sample_024", "sample_026", "sample_029",
                "sample_007", "sample_017", "sample_032"),
  condition = c("A", "A", "A",
                "B", "B", "B")
)
sample_metadata

### Conditions should be factors
# Statistical models in R need categorical variables stored as factors
# At the moment condition is a character column
class(sample_metadata$condition)
str(sample_metadata)

# We want condition a to be the reference
# ...so that later a positive fold change will mean "higher in b than a"

# Convert the condition column to a factor with a as the reference level
sample_metadata$condition <- factor(
  sample_metadata$condition,
  levels = c("A", "B")
)
# Check that the conversion worked
levels(sample_metadata$condition)
# Check the structure of the metadata table
str(sample_metadata)

### Row names

# Name rows of the metadata after the samples
rownames(sample_metadata) <- sample_metadata$sample_id

# Confirm the metadata table is correct after these changes
sample_metadata

### Finding the count files

# Locate the count files assigned for the assessment
# We need to be specific here as there are many files
# Could also replace with params above to be safe
condition_A_files <- c(
  "sample_024_counts.txt",
  "sample_026_counts.txt",
  "sample_029_counts.txt"
)

condition_B_files <- c(
  "sample_007_counts.txt",
  "sample_017_counts.txt",
  "sample_032_counts.txt"
)

# Define the file path for our files
# This is the only time we are using an absolute path
# This was called for in the assessment instructions
counts_dir <- "/home/shared/SLE777/A4_counts"

# Create vectors to save the files and their paths
count_files <- c(condition_A_files, condition_B_files)
count_paths <- file.path(counts_dir, count_files)

# Confirm that you can see the full file paths and file names
count_paths

# Safety checks to stop if not all six samples
stopifnot(length(count_files) == 6)
stopifnot(all(file.exists(count_paths)))

### Deriving sample names from file names

# Check file names again for what to strip
count_files

# Strip until we have sample identifiers only
sample_ids <- gsub("_counts\\.txt$", "", count_files)
sample_ids

# Check that every sample id derived from file name also appears in metadata table
# Stop the script if all results do not evaluate as TRUE
stopifnot(all(sample_ids %in% sample_metadata$sample_id))

####################################
# Merging the count files into one table
####################################

# Each count file is a two-column table: 
# ...a gene identifier, and the count for that one sample 
# To put six samples side by side
# ...we need to join these tables so that each row still refers to the same gene 

### Reading all six files with a loop
# The if else was added to the loop to create the counts_table object from scratch
# This deviates from the practical code as that included demonstration code
for (i in 1:length(count_paths)) {
  
  one_sample <- read.table(
    count_paths[i],
    header = FALSE,
    col.names = c("gene_id", sample_ids[i])
  )
  
  # Quick checks to reflect structure set earlier in code
  # Stop if the count table being processed does not have two columns
  # or if the first column is not character
  # or if the second column is not numeric
  stopifnot(
    ncol(one_sample) == 2,
    is.character(one_sample$gene_id),
    is.numeric(one_sample[[2]])
  )

  # Merge the new sample into the growing counts table
  # The first time through the loop the counts_table does not yet exist
  if (i == 1) {
    counts_table <- one_sample
  } else {
    counts_table <- merge(counts_table, one_sample, by = "gene_id")
  }
}

### Quick checks and guardrails for counts_table creation

# Confirm merged table contains one gene_id column plus six sample columns
stopifnot(ncol(counts_table) == length(sample_ids) + 1)

# Column names should be gene ID plus our sample IDs
expected_colnames <- c("gene_id", sample_ids)
stopifnot(all(colnames(counts_table) == expected_colnames))

# Check that the final table to see how many gene IDs and samples you have
# Remember that these are still inclusive of summary rows
message(
  "Merged count table created successfully: ",
  nrow(counts_table), " rows x ",
  ncol(counts_table), " columns."
)

####################################
# Separating the HTSeq-count summary rows
####################################

# Every count file has summary rows
# These were merged along with gene counts and will need to be separated

# Apply grepl to the wider table
is_summary <- grepl("^__", counts_table$gene_id)
sum(is_summary)

# Split table in two
htseq_summary <- counts_table[is_summary, ]
gene_table <- counts_table[!is_summary, ]

# Check that the split worked
dim(counts_table)

# Check that summary rows are in the summary table
dim(htseq_summary)

# Check that the gene table has no summary rows
dim(gene_table)

### Guardrails to be confident we have removed them

# Check that the summary table has five rows, one for each of the five HTSeq-count summary categories
stopifnot(nrow(htseq_summary) == 5)

# Check that the gene table has no summary rows
stopifnot(!any(grepl("^__", gene_table$gene_id)))

# And lastly, check that the gene table has the same number of rows as the original count files
# ...because we have not lost any genes in the merge
# The number of rows in the gene table should match the number of rows in any one of the original count files, minus the five summary rows
stopifnot(nrow(gene_table) == nrow(one_sample) - 5)

### These rows tell you how many reads were discarded because they fell outside any gene
# ...because they were ambiguous between overlapping genes
# or because their alignment quality was too low
# A sample with an unusually high __no_feature count deserves a closer look
htseq_summary

####################################
# Building the count matrix
####################################

# Need to remove gene_id so there is no text in table
# This is so that we can convert the table to a numeric matrix
count_matrix <- as.matrix(gene_table[, -1])
rownames(count_matrix) <- gene_table$gene_id

# Manually check the new object has the same dimensions minus one column
# Inspect visually the structure and first few rows
dim(count_matrix)
str(count_matrix)
head(count_matrix)

# And guardrailds for safety
stopifnot(
  nrow(count_matrix) == nrow(gene_table),
  ncol(count_matrix) == ncol(gene_table) - 1,
  all(rownames(count_matrix) == gene_table$gene_id),
  all(colnames(count_matrix) == sample_ids),
  is.matrix(count_matrix),
  is.numeric(count_matrix)
)

####################################
# Checking that the samples line up
####################################

# The columns of the count matrix and the rows of the metadata 
# ...must describe the same samples in the same order
# Nothing enforces this so we watch for silent errors

# First look at the two sets of names side by side
colnames(count_matrix)
sample_metadata$sample_id

# Now force the columns into the metadata order and check
count_matrix <- count_matrix[, sample_metadata$sample_id]
all(colnames(count_matrix) == sample_metadata$sample_id)

# Also get R to check for us
stopifnot(all(colnames(count_matrix) == sample_metadata$sample_id))

### Always reorder the matrix columns to match the metadata rows
# ...always assert that they match
# ...and never continue past a failed check
```

# 2. Normalisation and PCA

```{r pca}

```

# 3. Differential expression

```{r deseq2}

```

# 4. Significant genes

```{r significant}

```

# 5. Volcano plot and heatmap

```{r figures}

```

# 6. Annotation and GO enrichment

```{r enrichment}

```

# 7. Your gene of interest

```{r goi}

```

# 8. Interpretation

**Gene of interest (max 4 sentences):**

**PCA (max 3 sentences):**

**Enrichment (max 4 sentences):**

**Volcano caption (2 sentences):**

# 9. Session information

```{r session}
sessionInfo()
```
