---
title: "SLE777 Assessment 4: Reproducible Differential Expression Report"
author: "Ivy Craw, Murray Pung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params: 
  group_id: "24"
  # Six assigned count files exactly as they appear in our allocation row
  condition_a_files: ["sample_024_counts.txt","sample_026_counts.txt","sample_029_counts.txt"]
  condition_b_files: ["sample_007_counts.txt","sample_017_counts.txt","sample_032_counts.txt"]
  counts_dir: "/home/shared/SLE777/A4_counts"
  gene_of_interest: "YDL022W"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 6)
set.seed(777)
# Load the packages you need. Record versions with sessionInfo() at the end.
library(clusterProfiler)
library(DESeq2)
library(ggplot2)
# What's magrittr used for?
library(magrittr)
library(apeglm)
library(pheatmap)
library(org.Sc.sgd.db)
library(AnnotationDbi)

# Suggested library green lit by unit chair for dynamic labelling
# OPTIONAL
library(ggrepel)

# Create figures directory if it does not already exist
dir.create("figures", showWarnings = FALSE)
```

This report analyses the cohort assigned to group ``r params$group_id``,
gene of interest ``r params$gene_of_interest``.

# 1. Count matrix and metadata

```{r matrix}
####################################
# Build the count matrix
####################################

# Combine six HTSeq-count files into one gene six sample matrix
# Samples: condition A versus condition B, for assessment purposes

# To compare gene expression between conditions, we need all six samples side by side in one table
# The end product of this is a count matrix
# One row per gene and one column per sample and a read count in every cell
# We also build a sample metadata table
# These two objects are required for differential expression analysis (DESeq2)

### Descriptions of samples
# Counts provided for assessment purposes
# No additional information provided

####################################
# The sample metadata table
####################################

# Create sample metadata table with sample identifiers and condition
sample_metadata <- data.frame(
  sample_id = c("sample_024", "sample_026", "sample_029",
                "sample_007", "sample_017", "sample_032"),
  condition = c("A", "A", "A",
                "B", "B", "B")
)
sample_metadata

### Conditions should be factors
# Statistical models in R need categorical variables stored as factors
# At the moment condition is a character column
class(sample_metadata$condition)
str(sample_metadata)

# We want condition a to be the reference
# ...so that later a positive fold change will mean "higher in b than a"

# Convert the condition column to a factor with a as the reference level
sample_metadata$condition <- factor(
  sample_metadata$condition,
  levels = c("A", "B")
)
# Check that the conversion worked
levels(sample_metadata$condition)
# Check the structure of the metadata table
str(sample_metadata)

### Row names

# Name rows of the metadata after the samples
rownames(sample_metadata) <- sample_metadata$sample_id

# Confirm the metadata table is correct after these changes
sample_metadata

### Finding the count files

# Locate the count files assigned for the assessment
# We need to be specific here as there are many files
# Could also replace with params above to be safe
condition_A_files <- c(
  "sample_024_counts.txt",
  "sample_026_counts.txt",
  "sample_029_counts.txt"
)

condition_B_files <- c(
  "sample_007_counts.txt",
  "sample_017_counts.txt",
  "sample_032_counts.txt"
)

# Define the file path for our files
# This is the only time we are using an absolute path
# This was called for in the assessment instructions
counts_dir <- "/home/shared/SLE777/A4_counts"

# Create vectors to save the files and their paths
count_files <- c(condition_A_files, condition_B_files)
count_paths <- file.path(counts_dir, count_files)

# Confirm that you can see the full file paths and file names
count_paths

# Safety checks to stop if not all six samples
stopifnot(length(count_files) == 6)
stopifnot(all(file.exists(count_paths)))

### Deriving sample names from file names

# Check file names again for what to strip
count_files

# Strip until we have sample identifiers only
sample_ids <- gsub("_counts\\.txt$", "", count_files)
sample_ids

# Check that every sample id derived from file name also appears in metadata table
# Stop the script if all results do not evaluate as TRUE
stopifnot(all(sample_ids %in% sample_metadata$sample_id))

####################################
# Merging the count files into one table
####################################

# Each count file is a two-column table: 
# ...a gene identifier, and the count for that one sample 
# To put six samples side by side
# ...we need to join these tables so that each row still refers to the same gene 

### Reading all six files with a loop
# The if else was added to the loop to create the counts_table object from scratch
# This deviates from the practical code as that included demonstration code
for (i in 1:length(count_paths)) {
  
  one_sample <- read.table(
    count_paths[i],
    header = FALSE,
    col.names = c("gene_id", sample_ids[i])
  )
  
  # Quick checks to reflect structure set earlier in code
  # Stop if the count table being processed does not have two columns
  # or if the first column is not character
  # or if the second column is not numeric
  stopifnot(
    ncol(one_sample) == 2,
    is.character(one_sample$gene_id),
    is.numeric(one_sample[[2]])
  )

  # Merge the new sample into the growing counts table
  # The first time through the loop the counts_table does not yet exist
  if (i == 1) {
    counts_table <- one_sample
  } else {
    counts_table <- merge(counts_table, one_sample, by = "gene_id")
  }
}

### Quick checks and guardrails for counts_table creation

# Confirm merged table contains one gene_id column plus six sample columns
stopifnot(ncol(counts_table) == length(sample_ids) + 1)

# Column names should be gene ID plus our sample IDs
expected_colnames <- c("gene_id", sample_ids)
stopifnot(all(colnames(counts_table) == expected_colnames))

# Check that the final table to see how many gene IDs and samples you have
# Remember that these are still inclusive of summary rows
message(
  "Merged count table created successfully: ",
  nrow(counts_table), " rows x ",
  ncol(counts_table), " columns."
)

####################################
# Separating the HTSeq-count summary rows
####################################

# Every count file has summary rows
# These were merged along with gene counts and will need to be separated

# Apply grepl to the wider table
is_summary <- grepl("^__", counts_table$gene_id)
sum(is_summary)

# Split table in two
htseq_summary <- counts_table[is_summary, ]
gene_table <- counts_table[!is_summary, ]

# Check that the split worked
dim(counts_table)

# Check that summary rows are in the summary table
dim(htseq_summary)

# Check that the gene table has no summary rows
dim(gene_table)

### Guardrails to be confident we have removed them

# Check that the summary table has five rows, one for each of the five HTSeq-count summary categories
stopifnot(nrow(htseq_summary) == 5)

# Check that the gene table has no summary rows
stopifnot(!any(grepl("^__", gene_table$gene_id)))

# And lastly, check that the gene table has the same number of rows as the original count files
# ...because we have not lost any genes in the merge
# The number of rows in the gene table should match the number of rows in any one of the original count files, minus the five summary rows
stopifnot(nrow(gene_table) == nrow(one_sample) - 5)

### These rows tell you how many reads were discarded because they fell outside any gene
# ...because they were ambiguous between overlapping genes
# or because their alignment quality was too low
# A sample with an unusually high __no_feature count deserves a closer look
htseq_summary

####################################
# Building the count matrix
####################################

# Need to remove gene_id so there is no text in table
# This is so that we can convert the table to a numeric matrix
count_matrix <- as.matrix(gene_table[, -1])
rownames(count_matrix) <- gene_table$gene_id

# Manually check the new object has the same dimensions minus one column
# Inspect visually the structure and first few rows
dim(count_matrix)
str(count_matrix)
head(count_matrix)

# And guardrailds for safety
stopifnot(
  nrow(count_matrix) == nrow(gene_table),
  ncol(count_matrix) == ncol(gene_table) - 1,
  all(rownames(count_matrix) == gene_table$gene_id),
  all(colnames(count_matrix) == sample_ids),
  is.matrix(count_matrix),
  is.numeric(count_matrix)
)

####################################
# Checking that the samples line up
####################################

# The columns of the count matrix and the rows of the metadata 
# ...must describe the same samples in the same order
# Nothing enforces this so we watch for silent errors

# First look at the two sets of names side by side
colnames(count_matrix)
sample_metadata$sample_id

# Now force the columns into the metadata order and check
count_matrix <- count_matrix[, sample_metadata$sample_id]
all(colnames(count_matrix) == sample_metadata$sample_id)
identical(colnames(count_matrix), rownames(sample_metadata))

# Also get R to check for us
stopifnot(all(colnames(count_matrix) == sample_metadata$sample_id))
stopifnot(identical(colnames(count_matrix), rownames(sample_metadata)))

### Always reorder the matrix columns to match the metadata rows
# ...always assert that they match
# ...and never continue past a failed check

####################################
# Library sizes and read assignments
####################################

# The library size of a sample is the total number of reads assigned to genes
# Samples with more reads will show higher counts for every gene
# ...which is why we cannot compare raw counts between samples
# Normalising converts counts into a quantity that does not depend on how deeply the sample was sequenced

# Remember to use colSums
# A lib size is a total per sample, and samples are columns in our matrix
library_sizes <- colSums(count_matrix)
length(library_sizes)
library_sizes

# Scale library sizes to millions of reads and round to 2 decimal places
round(library_sizes / 1e6, 2)

### Summary of reads that were not assigned to any gene

# Summary rows we set aside earlier
# Convert to matrix, removing header, then add up each column
summary_matrix <- as.matrix(htseq_summary[, -1])
rownames(summary_matrix) <- htseq_summary$gene_id
summary_matrix

# Sum the rows of the summary matrix to get the total number of reads that were not assigned to any gene
unassigned <- colSums(summary_matrix)
unassigned

### Calculate the proportion of reads assigned to genes

# First build a data frame
qc_table <- data.frame(
  sample_id = names(library_sizes),
  library_size = as.numeric(library_sizes),
  unassigned = as.numeric(unassigned[names(library_sizes)])
)
qc_table

# Add two more columns using existing columns
qc_table$total_reads <- qc_table$library_size + qc_table$unassigned
qc_table$percent_assigned <- 100 * qc_table$library_size / qc_table$total_reads
qc_table

# A healthy RNA-seq sample from a well-annotated genome such as yeast should assign a large majority of its reads to genes
# A sample that is markedly worse than the others is a candidate for exclusion
# Should a least be flagged in report

### Plotting the library sizes

# We need to merge condition alongside the library size for coloured bars
qc_table <- merge(qc_table, sample_metadata, by = "sample_id")
qc_table

# Fill aesthetics maps a column to bar colour
# theme_bw() function replaces the grey default background
# theme() call rotates the x-axis labels so that the long sample names do not overlap
ggplot(qc_table, aes(x = sample_id, y = library_size / 1e6, fill = condition)) +
  geom_col() +
  labs(
    title = "Library size per sample",
    x = "Sample",
    y = "Assigned reads (millions)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

### Plot percent_assigned for each sample as a bar chart, coloured by condition, with a sensible title and axis labels
ggplot(qc_table, aes(x = sample_id, y = percent_assigned, fill = condition)) +
  geom_col() +
  labs(
    title = "Percentage of reads assigned to genes",
    x = "Sample",
    y = "Reads assigned (%)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


####################################
# A first look across samples
####################################

# RNA-seq counts often have a skewed distribution
# Many genes have low counts while a smaller number have very high counts
# Log transformation makes the distributions easier to compare across samples

# Draw histogram of counts across samples
# The las = 2 argument turns the sample labels sideways so they fit
# Logarithm cannot take zero so we add 1 across whole matrix
boxplot(
  log10(count_matrix + 1),
  las = 2,
  ylab = "log10(count + 1)",
  main = "Distribution of gene counts per sample"
)

####################################
# Filtering genes with no expression
####################################

# Many genes are not expressed at all in this experiment
# They carry no information, they slow the analysis down
# ...and they cost us statistical power when we correct for multiple testing in Week 9

### Count how many genes have zero reads across every sample

# Create new var for total n of rows count as total number of genes
gene_totals <- rowSums(count_matrix)

### Remove the genes with zero reads across samples

# Create new variable to store filtered new matrix
# Keep all columns
count_matrix_filtered <- count_matrix[gene_totals > 0, ]

# Calculate summary statistics of our genes across samples
dim(count_matrix)

# Summary statistics for only genes with reads
dim(count_matrix_filtered)

# Finally, check what working tables we've created thus far
ls()
```

# 2. Normalisation and PCA

To correct for sequencing depth, we divide each count by the sample's library size, before multiplying by one million. 

This is to make our counts more readable. 

Scaling like this, however, does not correct for gene length. For this reason, counts per million values cannot be compared between different genes. 
They can be compared for the When we are looking to do differential expression, scaling our counts per million can be useful to compare the *same* gene across samples. 

```{r pca}
##############################################
# Normalisation and exploring the experiment
##############################################

### Counts per million

# Each sample has its own library size
# Work through columns one at a time

# Copy count matrix so dimensions are the same
cpm <- count_matrix

# For each column i take col of counts
# divide it by that sample's lib size
# multiply by a million
# and put answer back in the count matrix copy cpm
for (i in 1:ncol(count_matrix)) {
  cpm[, i] <- count_matrix[, i] / library_sizes[i] * 1e6
}

# Check that we still have the same number genes and six samples
dim(cpm)
# Inspect results
head(cpm)
# and confirm that every column of CPM matrix sum is exactly one mill 1e+06
colSums(cpm)


### 6. Filtering low-count genes

# We cannot reliably test genes with almost no reads anywhere
# They inflate the multiple-testing correction we apply later
# costing us the power to detect genes of interest

# A widely used rule is to keep a gene if it has a CPM above 1 in at least as many samples as the smallest group
# Our smallest group has three replicates

keep <- rowSums(cpm > 1) >= 3
table(keep)

# The expression cpm > 1 gives a boolean matrix the same shape as cpm
# R treats TRUE as 1, so rowSums() of that matrix counts, for each gene, how many samples passed
# We then keep the genes where that count is at least three

# Apply the filter to both matrices
counts_filtered <- count_matrix[keep, ]
cpm_filtered <- cpm[keep, ]
dim(count_matrix)
dim(counts_filtered)

# Filtering through a logical vector before the comma, all columns kept
# Note that we keep the filtered raw counts as well
# DESeq2 next week wants raw counts, not CPM, because it models the counting process itself
# The CPM matrix is for exploration and plotting only

### 7. Log transformation

# Count data are strongly skewed
# A handful of genes carry enormous counts and dominate every plot
# Taking logarithms compresses that range
# As always, we add 1 first, because the log of zero is undefined

log_cpm <- log2(cpm_filtered + 1)

# Inspect the log CPM matrix
dim(log_cpm)
head(log_cpm)

# Earlier in the srcipt we used log10 to squash the y-axis of a boxplot
# Here we switch to log2, and the change is deliberate
# Base 2 is conventional in genomics because a difference of 1 on the log2 scale means a doubling of expression
# ...a difference of 2 means a fourfold change, and so on
# That interpretation carries directly into the log2 fold changes we calculate later in the analysis
# For simply squashing a plot's axis either base works; for talking about fold changes, base 2 is the one that lets you read the number

### 8. Did the normalisation work?

# We have claimed that normalisation makes samples comparable
# Do not take it on trust. Draw the distributions before and after and look.
# Because both matrices are numeric, base R gives one box per sample from a single line of code

boxplot(
  log2(counts_filtered + 1),
  las = 2,
  ylab = "log2(count + 1)",
  main = "Before normalisation"
)

boxplot(
  log_cpm,
  las = 2,
  ylab = "log2(CPM + 1)",
  main = "After normalisation"
)

##################################
# Principal component analysis ###
##################################

# Let's consider
# if the two stages really differ
# ...samples from the same stage should look more like each other than like samples from the other stage

# Our data have thousands of dimensions, one per gene, and we cannot draw that
# PCA finds a small number of new axes, called principal components, that capture as much of the variation as possible
# It lets us show six samples measured on thousands of genes on an ordinary scatter plot
# PCA needs parts per million

# Each component is built from a combination of many genes
# PC1 is the direction along which the samples differ most
# PC2 the next most, and so on
# If the experiment worked, PC1 usually separates the conditions

# 9.1 Transposing
# Risk of silent error here! 

# Transpose log-CPM matrix so samples are rows and genes are columns for prcomp()
# Run PCA to summarise the major patterns of variation between samples
pca <- prcomp(t(log_cpm))
dim(pca$x)
rownames(pca$x)

# The sample coordinates live in pca$x
# ...so if I ran colnames(pca$x) I get "PC1" "PC2" "PC3" "PC4" "PC5" "PC6"

# Thousands of gene-expression measurements per sample → PCA 
# → a few summary dimensions (PC1, PC2, etc.) describing the major differences between samples

# PC1 → captures the most variation
# PC2 → captures the second most
# PC3 → captures the third most
# PC4 → captures the fourth most
...
# Look at how much variation each component captures
percent_var <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

# Inspect variation explained by the first four (highest-ranked) PCs
percent_var[1:4]

# Breaking this down...
# prcomp() stores a standard deviation for each component in pca$sdev
# Variance is the square of standard deviation, so ^2
# Dividing each variance by the total gives each component's share
# Multiplying by 100 makes it a percentage

# Create a data frame containing each sample's coordinates
# on PC1 and PC2 (the two components explaining the most variation)
ca_df <- data.frame(
  sample_id = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2]
)

# Add sample metadata so PCA points can be identified by condition/stage
pca_df <- merge(pca_df, sample_metadata, by = "sample_id")
pca_df

# ...note, a biologically important treatment effect can still be relatively small and therefore not obvious in those first two PCs

### 10. Building the PCA figure with ggplot2
# The single most informative figure in a routine RNA-seq analysis
# So we create it slowly and thoughtfully

# 1. Data frame, first argument to ggplot()
# 2. Then aesthetic mapping, which column becomes x and y, colour
# ... we'll get an empty panel with correctly scaled axes
ggplot(pca_df, aes(x = PC1, y = PC2))

# 3. Now we give geom argument, what is actually drawn
# Six points
ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point()

# 4. Now we map the stage to colour and make points bigger
# Note, if the value comes from a column of your data, it goes inside aes()
# If you picked it yourself, it goes outside
# The same holds for colour, fill, size, shape and alpha
ggplot(pca_df, aes(x = PC1, y = PC2, colour = condition)) +
  geom_point(size = 4)

# 5. Set the colours explicitly for reporting
# Name them after the factor levels to the colour always follows the group
# Always use levels() e.g., pca_df$condition to check correct naming of data
ggplot(pca_df, aes(x = PC1, y = PC2, colour = condition)) +
  geom_point(size = 4) +
  scale_colour_manual(values = c("a" = "#66c2a5", "b" = "#fc8d62")) +
  theme_bw()

# 10.2 Labels and saving

# Set ggplot object for PCA plot
# paste0() glues strings together without a separator
p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, colour = condition)) +
  geom_point(size = 4) +
  scale_colour_manual(values = c("a" = "#66c2a5", "b" = "#fc8d62")) +
  labs(
    title = "PCA of normalised expression",
    x = paste0("PC1 (", percent_var[1], "% variance)"),
    y = paste0("PC2 (", percent_var[2], "% variance)"),
    colour = "Stage"
  ) +
  theme_bw()

# Draw object
p_pca

# Label the points
# I'm using ggrepel rather than manually setting using vjust
# OPTIONAL ADDITION confirmed by unit chair to be acceptable

p_pca_labels <- p_pca +
  geom_text_repel(
    aes(label = sample_id),
    size = 3,
    show.legend = FALSE
  )

# Display graph
p_pca_labels

# Save as png using ggsave()
# A resolution of 300 dpi is the usual minimum for a journal figure
ggsave(
  filename = "figures/pca_plot.png",
  plot = p_pca_labels,
  width = 6,
  height = 5,
  dpi = 300
)

# Say a b sample appeared among the a
# List three possible explanations ranging from a labelling mistake in the metadata to genuine biology
# How would we tell them apart?
# ...my answers are a) metadata labeling mistake in the a ggplot mapping or earlier
# b) one sample didn't ferment as expected
# c) sample might have a technical/sample qual issue, lower seq qual or unusual lib size
# ...contamination, poor alignment, etc. I'd go back to QC

# If PC1 separates your conditions and explains a large share of the variance
# ...the experiment has a strong effect and differential expression will find plenty of genes 
# If the conditions do not separate at all, be cautious about any gene list you produce afterwards
# Look at the data before you test it

```

# 3. Differential expression

```{r deseq2}

```

# 4. Significant genes

```{r significant}

```

# 5. Volcano plot and heatmap

```{r figures}

```

# 6. Annotation and GO enrichment

```{r enrichment}

```

# 7. Your gene of interest

```{r goi}

```

# 8. Interpretation

**Gene of interest (max 4 sentences):**

**PCA (max 3 sentences):**

**Enrichment (max 4 sentences):**

**Volcano caption (2 sentences):**

# 9. Session information

```{r session}
sessionInfo()
```
