---
title: "SLE777 Assessment 4: Reproducible Differential Expression Report"
author: "Ivy Craw, Murray Pung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params: 
  group_id: "24"
  # Six assigned count files exactly as they appear in our allocation row
  condition_a_files: ["sample_024_counts.txt","sample_026_counts.txt","sample_029_counts.txt"]
  condition_b_files: ["sample_007_counts.txt","sample_017_counts.txt","sample_032_counts.txt"]
  counts_dir: "/home/shared/SLE777/A4_counts"
  gene_of_interest: "YDL022W"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 6)
set.seed(777)
# Load the packages you need. Record versions with sessionInfo() at the end.
library(clusterProfiler)
library(DESeq2)
library(ggplot2)
# What's magrittr used for?
library(magrittr)
library(apeglm)
library(pheatmap)
library(org.Sc.sgd.db)
library(AnnotationDbi)
library(ggrepel)

# Create figures directory if it does not already exist
dir.create("figures", showWarnings = FALSE)
```

This report analyses the cohort assigned to group ``r params$group_id``,
gene of interest ``r params$gene_of_interest``. 

# 1. Count matrix and metadata

This section constructs the RNA-seq count matrix and sample metadata required for downstream analysis. The six assigned HTSeq-count files are read, merged into a single gene-by-sample matrix, and checked to ensure that sample names and metadata are correctly aligned. HTSeq summary rows are separated from gene counts, library sizes and read-assignment statistics are calculated for quality control, and visualisations of sequencing depth, read assignment rates, and count distributions are generated to assess sample quality before normalisation and differential expression analysis.

```{r matrix}
####################################
# Build the count matrix
####################################

# Combine six HTSeq-count files into one gene six sample matrix
# Samples: condition A versus condition B, for assessment purposes

# To compare gene expression between conditions
# we need all six samples side by side in one table
# The end product of this is a count matrix
# One row per gene and one column per sample and a read count in every cell
# We also build a sample metadata table
# These two objects are required for differential expression analysis (DESeq2)

### Descriptions of samples
# Counts provided for assessment purposes
# No additional information provided

####################################
# The sample metadata table
####################################

# Create sample metadata table with sample identifiers and condition
sample_metadata <- data.frame(
  sample_id = c("sample_024", "sample_026", "sample_029",
                "sample_007", "sample_017", "sample_032"),
  condition = c("A", "A", "A",
                "B", "B", "B")
)
sample_metadata

### Conditions should be factors
# Statistical models in R need categorical variables stored as factors
# At the moment condition is a character column
class(sample_metadata$condition)
str(sample_metadata)

# We want condition a to be the reference
# ...so that later a positive fold change will mean "higher in B than A"

# Convert the condition column to a factor with a as the reference level
sample_metadata$condition <- factor(
  sample_metadata$condition,
  levels = c("A", "B")
)

# Check that the conversion worked
levels(sample_metadata$condition)
# Check the structure of the metadata table
str(sample_metadata)

### Row names

# Name rows of the metadata after the samples
rownames(sample_metadata) <- sample_metadata$sample_id

# Confirm the metadata table is correct after these changes
sample_metadata

### Finding the count files

# Locate the count files assigned for the assessment
# We need to be specific here as there are many files
# Could also replace with params above to be safe
condition_A_files <- c(
  "sample_024_counts.txt",
  "sample_026_counts.txt",
  "sample_029_counts.txt"
)

condition_B_files <- c(
  "sample_007_counts.txt",
  "sample_017_counts.txt",
  "sample_032_counts.txt"
)

# Define the file path for our files
# This is the only time we are using an absolute path
# This was called for in the assessment instructions
#counts_dir <- "/home/shared/SLE777/A4_counts"
counts_dir <- "counts"

# Create vectors to save the files and their paths
count_files <- c(condition_A_files, condition_B_files)
count_paths <- file.path(counts_dir, count_files)

# Confirm that you can see the full file paths and file names
count_paths

# Safety checks to stop if not all six samples
stopifnot(length(count_files) == 6)
stopifnot(all(file.exists(count_paths)))

### Deriving sample names from file names

# Check file names again for what to strip
count_files

# Strip until we have sample identifiers only
sample_ids <- gsub("_counts\\.txt$", "", count_files)
sample_ids

# Check that every sample id derived from file name also appears in metadata
# Stop the script if all results do not evaluate as TRUE
stopifnot(all(sample_ids %in% sample_metadata$sample_id))

####################################
# Merging the count files into one table
####################################

# Join the six samples side by side
# to merge gene counts

### Reading all six files with a loop
# The if else was added to create the counts_table object from scratch
# This deviates from the practical code as that included demonstration code
for (i in 1:length(count_paths)) {
  
  one_sample <- read.table(
    count_paths[i],
    header = FALSE,
    col.names = c("gene_id", sample_ids[i])
  )
  
  # Quick checks to reflect structure set earlier in code
  # Stop if the count table being processed does not have two columns
  # or if the first column is not character
  # or if the second column is not numeric
  stopifnot(
    ncol(one_sample) == 2,
    is.character(one_sample$gene_id),
    is.numeric(one_sample[[2]])
  )

  # Merge the new sample into the growing counts table
  # The first time through the loop the counts_table does not yet exist
  if (i == 1) {
    counts_table <- one_sample
  } else {
    counts_table <- merge(counts_table, one_sample, by = "gene_id")
  }
}

### Quick checks and guardrails for counts_table creation

# Confirm merged table contains one gene_id column plus six sample columns
stopifnot(ncol(counts_table) == length(sample_ids) + 1)

# Column names should be gene ID plus our sample IDs
expected_colnames <- c("gene_id", sample_ids)
stopifnot(all(colnames(counts_table) == expected_colnames))

# Check that the final table to see how many gene IDs and samples you have
# Remember that these are still inclusive of summary rows
message(
  "Merged count table created successfully: ",
  nrow(counts_table), " rows x ",
  ncol(counts_table), " columns."
)

####################################
# Separating the HTSeq-count summary rows
####################################

# Every count file has summary rows
# These were merged along with gene counts and will need to be separated

# Apply grepl to the wider table
is_summary <- grepl("^__", counts_table$gene_id)
sum(is_summary)

# Split table in two
htseq_summary <- counts_table[is_summary, ]
gene_table <- counts_table[!is_summary, ]

# Check that the split worked
dim(counts_table)

# Check that summary rows are in the summary table
dim(htseq_summary)

# Check that the gene table has no summary rows
dim(gene_table)

### Guardrails to be confident we have removed them

# Check that the summary table has five rows
# one for each of the five HTSeq-count summary categories
stopifnot(nrow(htseq_summary) == 5)

# Check that the gene table has no summary rows
stopifnot(!any(grepl("^__", gene_table$gene_id)))

# And lastly, check that the gene table has the same number of rows as the original count files
# ...because we have not lost any genes in the merge
# The number of rows in the gene table 
# should match the number of rows in any one of the original count files
# minus the five summary rows
stopifnot(nrow(gene_table) == nrow(one_sample) - 5)

### These rows tell you how many reads were discarded because they fell outside any gene
# ...because they were ambiguous between overlapping genes
# or because their alignment quality was too low
# A sample with an unusually high __no_feature count deserves a closer look
htseq_summary

# Samples in condition B have higher absolute no_feature counts
# Particularly sample 007
# We'll keep an eye on library sizes below

####################################
# Building the count matrix
####################################

# Need to remove gene_id so there is no text in table
# This is so that we can convert the table to a numeric matrix
count_matrix <- as.matrix(gene_table[, -1])
rownames(count_matrix) <- gene_table$gene_id

# Manually check the new object has the same dimensions minus one column
# Inspect visually the structure and first few rows
dim(count_matrix)
str(count_matrix)
head(count_matrix)

# And guardrailds for safety
stopifnot(
  nrow(count_matrix) == nrow(gene_table),
  ncol(count_matrix) == ncol(gene_table) - 1,
  all(rownames(count_matrix) == gene_table$gene_id),
  all(colnames(count_matrix) == sample_ids),
  is.matrix(count_matrix),
  is.numeric(count_matrix)
)

####################################
# Checking that the samples line up
####################################

# The columns of the count matrix and the rows of the metadata 
# ...must describe the same samples in the same order
# Nothing enforces this so we watch for silent errors

# First look at the two sets of names side by side
colnames(count_matrix)
sample_metadata$sample_id

# Now force the columns into the metadata order and check
count_matrix <- count_matrix[, sample_metadata$sample_id]
all(colnames(count_matrix) == sample_metadata$sample_id)
identical(colnames(count_matrix), rownames(sample_metadata))

# Also get R to check for us
stopifnot(all(colnames(count_matrix) == sample_metadata$sample_id))
stopifnot(identical(colnames(count_matrix), rownames(sample_metadata)))

### Always reorder the matrix columns to match the metadata rows
# ...always assert that they match
# ...and never continue past a failed check

####################################
# Library sizes and read assignments
####################################

# The library size of a sample is the total number of reads assigned to genes
# Samples with more reads will show higher counts for every gene
# ...which is why we cannot compare raw counts between samples
# Normalising converts counts into a quantity that does not depend 
# ...on how deeply the sample was sequenced

# A lib size is a total per sample, and samples are columns in our matrix
library_sizes <- colSums(count_matrix)
# # Confirming we have our six samples
length(library_sizes)
library_sizes

# Scale library sizes to millions of reads and round to 2 decimal places
round(library_sizes / 1e6, 2)

# We can see that condition B samples have more reads assigned to genes
# likely consistent with greater sequencing depth

### Summary of reads that were not assigned to any gene

# Summary rows we set aside earlier
# Convert to matrix, removing header, then add up each column
summary_matrix <- as.matrix(htseq_summary[, -1])
rownames(summary_matrix) <- htseq_summary$gene_id
summary_matrix

# Sum the rows of the summary matrix to get the total number of reads that were not assigned to any gene
unassigned <- colSums(summary_matrix)
unassigned

### Calculate the proportion of reads assigned to genes

# First build a data frame
qc_table <- data.frame(
  sample_id = names(library_sizes),
  library_size = as.numeric(library_sizes),
  unassigned = as.numeric(unassigned[names(library_sizes)])
)
qc_table

# Add two more columns using existing columns
qc_table$total_reads <- qc_table$library_size + qc_table$unassigned
qc_table$percent_assigned <- 100 * qc_table$library_size / qc_table$total_reads

# Define sample order for plots so samples are grouped by condition
plot_sample_order <- sample_metadata$sample_id[
  order(sample_metadata$condition, sample_metadata$sample_id)
]

# Review table so far
qc_table

# A healthy RNA-seq sample from a well-annotated genome 
# should assign a large majority of its reads to genes
# A sample that is markedly worse than the others is a candidate for exclusion

# Let's add condition to investigate more closely and for bar colour
qc_table <- merge(qc_table, sample_metadata, by = "sample_id")

# Preserve the chosen sample order in ggplot
qc_table$sample_id <- factor(
  qc_table$sample_id,
  levels = plot_sample_order
)

# Review again and plot
qc_table

ggplot(qc_table, aes(x = sample_id, y = library_size / 1e6, fill = condition)) +
  geom_col() +
  labs(
    title = "Library size per sample",
    x = "Sample",
    y = "Assigned reads (millions)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# We can observe that condition B has more reads assigned to genes
# It could flag a potential issue, it may suggest corruption of data
# It may indicate a batch effect occurring or other issue

# Ideally we'd bring attention upstream, biologist or project manager who performed the sequencing
# They may say no batch effects, and it was due to reason x or otherwise
# If Condition A samples belong to a single batch, and Condition B samples from a different batch 
# then finding significance between the condition could just mean that batch effects are significant

# If there is a batch effect and it is correlated with condition
# then the experiment cannot be saved because we cannot adjust the model for batch effects
# Condition B samples had larger library sizes than condition A samples, indicating higher sequencing depth. Although this difference could introduce technical bias in raw counts, normalization was applied (presented in the PCA & Normalisation section below) to account for variation in library size, and all samples showed comparable read assignment rates (74.5-78.7%). No obvious outlier samples were identified based on sequencing depth alone.

### Plot percent_assigned for each sample as a bar chart, coloured by condition, with a sensible title and axis labels
ggplot(qc_table, aes(x = sample_id, y = percent_assigned, fill = condition)) +
  geom_col() +
  labs(
    title = "Percentage of reads assigned to genes",
    x = "Sample",
    y = "Reads assigned (%)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Condition A samples have slightly lower assignment rates,
# particularly sample_026, while condition B is more consistent at ~78%
# This could warrant further QC investigation upstream
# to rule out any potential batch effect or technical issues
# or provide legitimate technical or biological explanation 

# Condition B also has larger assigned libraries
# Its slightly higher assignment rates suggest that poorer assignment
# in condition A does not fully explain the difference in library sizes

# Note, percent_assigned is calculated from total_reads in our QC table
# These are reads accounted for by HTSeq, not necessarily all reads
# originally generated by the sequencer
# Confirming sequencing depth would require upstream FASTQ or sequencing QC data

#######################################
# A first look at counts across samples
#######################################

# RNA-seq counts have a skewed distribution
# Many genes have low counts while a smaller number have very high counts
# Log transformation makes the distributions easier to compare across samples

# Using the plot_sample_order var again to group samples
# The las = 2 argument turns the sample labels sideways so they fit
# Logarithm cannot take zero so we add 1 across whole matrix
boxplot(
  log10(count_matrix[, plot_sample_order] + 1),
  las = 2,
  ylab = "log10(count + 1)",
  main = "Distribution of gene counts per sample"
)

# The boxplot shows that condition B generally has higher raw gene counts than condition A
# although the distributions overlap considerably 
# All samples contain highly expressed genes

### Let's check the numbers

# Compare the distribution of gene counts across our six samples
# Let's see the numbers behind our boxplot
count_summary <- data.frame(
  sample_id = colnames(count_matrix),
  median_count = apply(count_matrix, 2, median),
  mean_count = round(colMeans(count_matrix), 1),
  max_count = apply(count_matrix, 2, max)
)

# Group samples by condition
count_summary <- merge(count_summary, sample_metadata, by = "sample_id")
count_summary <- count_summary[
  order(count_summary$condition, count_summary$sample_id),
]

# Show summary table
count_summary

```

The means are considerably higher than the medians, reflecting a skewed distribution with a smaller number of highly expressed genes as we saw in the boxplot. These are raw counts, however, so differences between our conditions can reflect library size rather than biological expression.

In a real-world study, we would investigate the differences in library sizes, proportions of genes assigned, and high no-feature reads with the biologist or sequencing team to determine whether they reflect sequencing depth, batch effects or another technical factor. 

If all condition A samples belonged to one batch and all condition B samples to another, batch and condition would be perfectly confounded. 

Any significant differences between conditions could reflect batch effects rather than biology, and we would not be able to adjust the model to separate the two.


# 2. Normalisation and PCA

To help correct for these differences in library size/sequencing depth, we divide each count by the sample's library size, before multiplying by one million. 

This is to make our counts more readable; we use CPM for visualisation and to perform our Principle Components Analysis.

Scaling our counts per million can be useful to compare the *same* gene across samples. 

It does not, however, correct for gene length. For this reason, counts per million values cannot be compared between different genes. 

For differential expression testing, DESeq2 uses the filtered raw counts and performs its own normalisation.
This section prepares the count data for downstream analysis by removing unexpressed and low-count genes, normalising counts to counts-per-million (CPM), and applying a log₂ transformation to stabilise the variance. Principal component analysis (PCA) is then performed on the normalised expression data to identify the major sources of variation between samples, quantify the proportion of variance explained by each component, and assess whether samples cluster according to experimental condition. The resulting PCA plot provides an overview of global expression patterns and serves as an important quality-control check before differential expression analysis.

```{r pca}
####################################
# Filtering genes with no expression
####################################

# Many genes may not be expressed at all in this experiment
# They carry no information, they slow the analysis down
# ...and they cost us statistical power when we correct for multiple testing

### Count how many genes have zero reads across every sample

# Create new var for total n of rows count as total number of genes
gene_totals <- rowSums(count_matrix)

### Remove the genes with zero reads across samples

# Create new variable to store filtered new matrix
# Keep all columns
count_matrix_filtered <- count_matrix[gene_totals > 0, ]

# Check dimensions before filtering
dim(count_matrix)

# and after filtering
dim(count_matrix_filtered)

# Quick check on how many we removed
sum(gene_totals == 0)

#############################
# Normalisation and exploring
#############################

### Counts per million

# Copy count matrix so dimensions are the same
cpm <- count_matrix

# For each column i take col of counts
# divide it by that sample's lib size
# multiply by a million
# and put answer back in the count matrix copy cpm
for (i in 1:ncol(count_matrix)) {
  cpm[, i] <- count_matrix[, i] / library_sizes[i] * 1e6
}

# Check that we still have the same number genes and six samples
dim(cpm)
# Inspect results
head(cpm)
# and confirm that every column of CPM matrix sum is one mill 1e+06
colSums(cpm)

### Filtering low-count genes

# We cannot reliably test genes with almost no reads anywhere
# They inflate the multiple-testing correction we apply later
# costing us the power to detect genes of interest

# A widely used rule is to keep a gene if it has a CPM above 1 in at least as many samples as the smallest group
# Our smallest group has three replicates
# table() will show us boolean TRUE FALSE values for how many genes passed this filter

keep <- rowSums(cpm > 1) >= 3
table(keep)

# Apply the filter to both matrices and retain all columns
# Check how many genes were dropped match the filter above
counts_filtered <- count_matrix[keep, ]
cpm_filtered <- cpm[keep, ]
dim(count_matrix)
dim(counts_filtered)

# Note that we keep the filtered raw counts as well
# DESeq2 below wants raw counts, not CPM, because it models the counting process itself
# The CPM matrix is for exploration and plotting only

### Log transformation

# Count data are strongly skewed as we have just observed
# A handful of genes carry enormous counts and dominate every plot
# Taking logarithms compresses that range
# As always, we add 1 first, because the log of zero is undefined

log_cpm <- log2(cpm_filtered + 1)

# Inspect the log CPM matrix
dim(log_cpm)
head(log_cpm)

# Earlier in the script we used log10 to display raw counts
# Here we use log2, which is conventional in genomics
# because a difference of 1 on the log2 scale means a doubling of expression
# For simply squashing a plot's axis either base works
# for talking about fold changes, base 2 is the one that lets you read the number

### Did the normalisation work?

# Draw the distributions before and after to manually check that normalisation has worked
# Check whether accounting for differences in sequencing depth and library composition
# reduces differences between samples

# Before normalisation
boxplot(
  log2(counts_filtered + 1),
  las = 2,
  ylab = "log2(count + 1)",
  main = "Before normalisation"
)

# After normalisation
boxplot(
  log_cpm,
  las = 2,
  ylab = "log2(CPM + 1)",
  main = "After normalisation"
)
# The normalisation appears to have worked - the differences caused by unequal sequencing depth no longer seem to dominate the data after normalization.


# Looks like the differences have reduced appropriately after normalisation
# Enough for us to compare gene expression
# Not expected to align perfectly
# as genuine biological variation should remain

##################################
# Principal component analysis ###
##################################

# Let's consider
# if the two stages really differ
# ...samples from the same stage should look more like each other than like samples from the other stage

# Our data have thousands of dimensions, one per gene, and we cannot draw that
# PCA finds a small number of new axes, principal components, that capture as much of the variation as possible
# It lets us show six samples measured on thousands of genes on an ordinary scatter plot
# PCA needs parts per million

# Each component is built from a combination of many genes
# PC1 is the direction along which the samples differ most
# PC2 the next most, and so on
# If the experiment worked, PC1 usually separates the conditions

### Transposing

# prcomp() expects observations x variables
# We have genes x samples, but our variables are genes
# So we need to switch! That's what t() does
# Run PCA to summarise the major patterns of variation between samples
pca <- prcomp(t(log_cpm))

# We should have the same number of PCA components as we do samples
dim(pca$x)
# Check that our row names have been switched to samples
rownames(pca$x)

# Add stop checks so R can check for us too
stopifnot(
  nrow(pca$x) == ncol(log_cpm),
  all(rownames(pca$x) == colnames(log_cpm))
)

# The sample coordinates live in pca$x

# Thousands of gene-expression measurements per sample → PCA 
# → a few summary dimensions (PC1, PC2, etc.) describing the major differences between samples

# PC1 → captures the most variation
# PC2 → captures the second most
# PC3 → captures the third most
# PC4 → captures the fourth most

# Create a data frame containing each sample's coordinates
# on PC1 and PC2 (the two components explaining the most variation)
pca_df <- data.frame(
  sample_id = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2]
)

# Add sample metadata so PCA points can be identified by condition/stage
pca_df <- merge(pca_df, sample_metadata, by = "sample_id")
pca_df

# ...note, a biologically important treatment effect can still be relatively small
# and therefore not obvious in those first two PCs

# Look at how much percent variation each component captures
# prcomp() stores a standard deviation for each component in pca$sdev
# Variance is the square of standard deviation, so ^2
# Dividing each variance by the total gives each component's share
# Multiplying by 100 makes it a percentage
percent_var <- round(
  100 * pca$sdev^2 / sum(pca$sdev^2),
  1
)

# Inspect variation explained by the first four (highest-ranked) PCs
percent_var[1:4]

### Building the PCA figure with ggplot2

## Data frame
# We map to our PCA dataframe pca_df

## Aesthetic mapping
# Map our top PCA points to x and y, and condition to colour

## Geom
# We set the size of points with geom_point and draw
# We label points with sample_id using ggrepel rather than manually setting using vjust

### scale_colour_manual
# Name them after the factor levels to the colour always follows the group
# Always use levels() e.g., pca_df$condition to check correct naming of data
pca_df$condition

## Labels
# Set ggplot object for PCA plot
# paste0() glues strings together without a separator

## Finally, theme_bw() just gives us a classic background colour scheme

# Draw the plot!
p_pca <- ggplot(
  pca_df,
  aes(x = PC1, y = PC2, colour = condition)
) +
  geom_point(size = 4) +
  geom_text_repel(
    aes(label = sample_id),
    size = 3,
    show.legend = FALSE
  ) +
  scale_colour_manual(
    values = c("A" = "#66c2a5", "B" = "#fc8d62")
  ) +
  labs(
    title = "PCA of normalised expression",
    x = paste0("PC1 (", percent_var[1], "% variance)"),
    y = paste0("PC2 (", percent_var[2], "% variance)"),
    colour = "Condition"
  ) +
  theme_bw()

# Draw object
p_pca

## IMPORTANT! 
# If the labels are confusing, try running that block again
# ggrepel will dynamically label the points so a reset will often fix funny positioning

# Save as png using ggsave()
# A resolution of 300 dpi is the usual minimum for a journal figure
ggsave(
  filename = "figures/pca_plot.png",
  plot = p_pca,
  width = 6,
  height = 5,
  dpi = 300
)

```

The PCA of normalised expression data showed clear separation of conditions A and B along PC1, which explained 53.7% of the total variance. All samples clustered according to their experimental condition along PC1, suggesting that condition is the primary source of variation in the dataset rather than differences in sequencing depth. Sample_026 exhibited substantial variation along PC2 (19.5% of variance) relative to the other A replicates, indicating additional within-group variability. However, sample_026 remained grouped with condition A on PC1 and did not cluster with the B samples, suggesting that it represents biological or technical variation within condition A rather than a failure of condition separation.

# 3. Differential expression

This section performs differential expression analysis using DESeq2 to identify genes whose expression differs between conditions A and B. Raw count data are modelled using a negative binomial framework, accounting for differences in sequencing depth through size-factor normalisation and estimating gene-specific dispersion to quantify biological variability. Differential expression statistics are then calculated for the contrast of condition B versus condition A, and log₂ fold changes are shrunk using the apeglm method to obtain more stable effect-size estimates. Diagnostic plots, including dispersion estimates, p-value distributions, and MA plots, are used to assess model performance and the effect of fold-change shrinkage.

```{r deseq2}
# let's check that the count and metadata are correctly aligned
dim(counts_filtered)
head(counts_filtered)
sample_metadata
all(colnames(counts_filtered) == sample_metadata$sample_id)
stopifnot(all(colnames(counts_filtered) == sample_metadata$sample_id))

# DEA is performed on the filtered raw counts, rather than the CPM data
# Build the DESeqDataSet
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered, # countData is the gene by sample matrix of raw counts
  colData = sample_metadata,   # colData is the metadata
  design = ~ condition)        
# model each gene's expression as a function of the condition column in the metadata (other variables can be used)
# If our samples had been prepared in two batches, and we knew which, we would write design = ~ batch + condition
# The variable of interest (condition) goes last in the formula. Everything before it is a nuisance variable to be adjusted for

# DESeq() is not in the unit material, but I can't get the resultsNames output correct unless I use it
dds <- DESeq(dds)
# check that dds correctly specifies condition
levels(dds$condition)
resultsNames(dds)

# DESeq2 computes a size factor per sample
# A size factor near 1 means the sample is close to typical. Above 1 means it was sequenced more deeply than average, and DESeq2 will scale it down accordingly.
dds <- estimateSizeFactors(dds)
sizeFactors(dds)

library_sizes <- colSums(counts_filtered)
cor_test <- cor.test(
  library_sizes / mean(library_sizes),
  sizeFactors(dds),
  method = "pearson")$estimate

data.frame(
  sample = colnames(dds),
  libsize_norm = round(library_sizes / mean(library_sizes), 3),
  sizefactor = round(sizeFactors(dds), 3))
paste0("Pearson's Correlation Coefficient estimate : ", round(cor_test,3))
# The correlation close to 1 indicates that DESeq2 size factors are largely tracking sequencing depth. When the correlation is substantially lower, it suggests that composition bias is present and DESeq2's median-ratio normalization is adjusting for more than library size.

# Dispersion and Shrinkage
# Variance in gene count between replicates is expected. This is referred to as dispersion, and the DEA pipeline 
# controls for the expected dispersion using the estimateDispersions function
# the final count is essentially a compromise between the gene specific estimate and the overall fitted trend
# the blue points that are notably well above the fitted line genes having between-sample variation much larger than expected
dds <- estimateDispersions(dds)
plotDispEsts(dds)

# Save dispersion plot to figures folder
png("figures/dispersion_plot.png", width = 1800, height = 1500, res = 300)
plotDispEsts(dds)
dev.off()

# Although the DEA steps above were performed manually, 
# the DESeq() function performs size factor estimation, dispersion estimation and model fitting in one call.
dds <- DESeq(dds)

res <- results(dds, contrast = c("condition", "B", "A"))
# A positive log fold change means higher in the B condition.

mcols(res)$description
summary(res)
# more genes are down regulated, zero outliers, 2% with low count after low count filter

# diagnostic histogram of p-values
res_raw_df <- as.data.frame(res)

ggplot(res_raw_df, aes(x = pvalue)) +
  geom_histogram(binwidth = 0.02, boundary = 0) +
  labs(title = "Distribution of raw p-values", x = "p-value", y = "Number of genes") +
  theme_bw()
# the spike near zero and flat towards one suggest the model is correctly specified

par(mfrow = c(1, 2)) # 1 row, 2 columns
plotMA(res, ylim = c(-6, 6), main = "Before shrinkage")
# The funnel shape to the left is created by weakly expressed genes fan out to extreme fold changes

# extract the shrunked counts for us to explore the significant genes.
res_shrunk <- lfcShrink(dds, coef = "condition_B_vs_A", type = "apeglm")
plotMA(res_shrunk, ylim = c(-6, 6), main = "After shrinkage")
#The funnel collapses. The strong, well-measured genes remain out at the edges
par(mfrow = c(1, 1)) # 1 row, 1 column

```

# 4. Significant genes

Here we are identifying the significant genes at padj < 0.05 and an absolute log2 fold change above 1, and splitting them by direction.

We rely on the following data objects returned by the DESeq2 analysis performed above:

- `dds` — not strictly needed for the basic filtering of significant genes, but useful downstream for plots/checks.
- `res` - the original, unshrunken DESeq2 B-vs-A results; useful for checking/testing and downstream analysis. 
- `res_shrunk` — shrunken results, which are used to construct our master data frame for identifying significant genes. 
- The correct coefficient used in lfcShrink() — it should come from resultsNames(dds) explicitly. 

```{r significant}
###################################
# Identifying significant genes ###
###################################

# Use the shrunken fold changes for ranking genes, for plotting, and for deciding which genes are biologically interesting
# Use the unshrunken p-values and adjusted p-values for deciding which genes are statistically significant
# Shrinkage changes the effect sizes, not the significance

# Turn the shrunken results into a plain data frame
# The third line reorders the columns so that gene_id comes first
res_df <- as.data.frame(res_shrunk)
res_df$gene_id <- rownames(res_shrunk)
res_df <- res_df[, c("gene_id", "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj")]
head(res_df)

# Remove the genes DESeq2 could not test, then sort by adjusted p-value
# !is.na() keeps the rows that have a real adjusted p-value
res_df <- res_df[!is.na(res_df$padj), ]
res_df <- res_df[order(res_df$padj), ]
head(res_df, 10)

# Apply both thresholds
# Statistical significance and a meaningful effect size are different requirements
# A good gene list demands both
# The abs() function takes the absolute value
# so this keeps genes that at least doubled in either direction
# Both conditions must hold
sig_genes <- res_df[res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1, ]

# Inspect the number of genes you get and the first few
nrow(sig_genes)
head(sig_genes)

# Split them by direction
# Remember that positive means higher in condition B, because A is the reference
up_in_b <- sig_genes[sig_genes$log2FoldChange > 0, ]
down_in_b <- sig_genes[sig_genes$log2FoldChange < 0, ]
nrow(up_in_b)
nrow(down_in_b)

# How many genes did DESeq2 leave untested
# so how many have padj set to NA?
sum(is.na(res$padj))

### Checking the result with our own eyes

# Take the top gene and plot its normalised counts in each sample
top_gene <- res_df$gene_id[1]
top_gene
plotCounts(dds, gene = top_gene, intgroup = "condition")

# The three A replicates should sit apart from the three B
# If they overlap, look at metadata closely to rule out labelling error

### Building the same figures in ggplot2 for a publishable version

### Calculate the top three genes

# This code is adapted from the unit content with suggestions from CoPilot
# Using index range from 1 to 3 top genes
# Because res_df is sorted by padj, indexing gives us the most significant genes
top_three_genes <- res_df$gene_id[1:3]
top_three_genes

# Extract normalised counts for the three most significant genes
# lapply(top_three_genes, function(gene) { runs function on each top gene
# gives us a list of three data frames so do.call(rbind, ...) stacks vertically
# it then runs the loop of plotcounts to get norm counts on each gene
top_three_counts <- do.call(
  rbind,
  lapply(top_three_genes, function(gene) {
    
    gene_norm <- plotCounts(
      dds,
      gene = gene,
      intgroup = "condition",
      returnData = TRUE
    )
    
    # Add sample IDs from the row names
    gene_norm$sample_id <- rownames(gene_norm)
    
    # Record which gene these counts belong to
    gene_norm$gene <- gene
    
    return(gene_norm)
  })
)

# Set gene factor levels to preserve significance order
top_three_counts$gene <- factor(
  top_three_counts$gene,
  levels = top_three_genes
)

# Plot all three genes
top_three_sig_genes <- ggplot(top_three_counts,
       aes(x = condition, y = count, colour = condition)) +
  geom_point(size = 4) +
  scale_y_log10() +
  scale_colour_manual(
    values = c("A" = "#66c2a5", "B" = "#fc8d62")
  ) +
  facet_wrap(~ gene) +
  labs(
    x = "Condition",
    y = "Normalised count"
  ) +
  theme_bw()

# OPTIONAL to include sample ids inside ggplot
#   geom_text_repel(aes(label = sample_id), size = 3, show.legend = FALSE) +
# But I think it's too squished once rendered

# Draw object
top_three_sig_genes

# OPTIONAL saving
ggsave(
  filename = "figures/top_three_sig_genes.png",
  plot = top_three_sig_genes,
  width = 6,
  height = 5,
  dpi = 300
)
```

# 5. Volcano plot and heatmap

This section identifies significant differentially expressed genes and visualises the results using a volcano plot and heatmap. Genes are classified as up-regulated in condition B, down-regulated in condition B, or not significant using thresholds of adjusted p-value < 0.05 and absolute log₂ fold change > 1. A volcano plot is generated to display both statistical significance and effect size, while a heatmap of significant genes is used to assess expression patterns, sample clustering, and consistency among biological replicates.

```{r figures}
res_df$neg_log10_padj <- -log10(res_df$padj)
head(res_df[, c("gene_id", "log2FoldChange", "padj", "neg_log10_padj")])

# Add a column recording each gene's status. We use the same thresholds as last week, and we name the groups after the stages we actually compared.
# "Up in B" is not a stylistic choice. Because B is the reference level, a positive fold change means the gene is higher in B than A, and that is the only claim these labels are entitled to make.
res_df$status <- "Not significant"
res_df$status[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up in B"
res_df$status[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down in B"
table(res_df$status)

top10 <- res_df[order(res_df$padj), ][1:10, ]

p_volcano <- ggplot(
  res_df,
  aes(
    x = log2FoldChange,
    y = neg_log10_padj,
    colour = status
  )
) +
  geom_point(
    alpha = 0.6,
    size = 1.5
  ) +
  scale_colour_manual(
    values = c(
      "Up in B" = "#d73027",
      "Down in B" = "#4575b4",
      "Not significant" = "grey70"
    )
  ) +
  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed",
    colour = "grey40"
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    colour = "grey40"
  ) +
  labs(
    title = "Differential expression: B versus A",
    x = "log2 fold change (B / A)",
    y = "-log10 adjusted p-value",
    colour = ""
  ) +
  theme_bw() +
  geom_text_repel(
     data = top10,
     aes(label = gene_id),
     size = 2.5,
     colour = "black")

p_volcano



# Heatmap
# Heatmaps answer - do the replicates within a group behave alike, and do the two groups (~condition) look different?
# We plot only the significant genes, because plotting several thousand rows would show nothing but noise.
# Pull out the significant gene identifiers, then subset the normalised expression matrix.

sig_ids <- res_df$gene_id[res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1]
sig_ids <- sig_ids[!is.na(sig_ids)]
length(sig_ids)

heat_matrix <- log_cpm[rownames(log_cpm) %in% sig_ids, ]
dim(heat_matrix)

# Build an annotation data frame so the heatmap can show which sample belongs to which stage.

annotation_col <- data.frame(Stage = sample_metadata$condition)
rownames(annotation_col) <- sample_metadata$sample_id

pheatmap(
  heat_matrix,
  scale = "row",
  show_rownames = FALSE,
  annotation_col = annotation_col,
  main = "Differentially expressed genes: B versus A")
# We see clear differential between the A vs B samples, and reasonable homogeneity between the samples within each Stage (A vs B) category
```

# 6. Annotation and GO enrichment

This section annotates differentially expressed genes using the yeast annotation database and performs Gene Ontology (GO) over-representation analysis to identify biological processes associated with the transcriptional response. Significant genes are mapped from systematic ORF identifiers to common gene names and functional descriptions, then compared against the tested gene universe using GO Biological Process enrichment analysis. Enrichment is assessed for all significant genes combined and separately for genes up-regulated and down-regulated in condition B. The results provide biological context for the differential expression analysis by identifying cellular processes that are disproportionately represented among the affected genes.

```{r enrichment}
##################################
### Annotating the gene names ####
##################################

# Bioconductor supplies annotation packages that translate common name from systematic
# I.e., FMP16 vs YDR070C

# the mapIds() function looks up one identifier type and returns another
# keys are our gene IDs
# keytype will be ORF for yeast systematic names
# we want the common names returned by column
# multiVals = "first" handles the genes that map to more than one name by taking the first
res_df$gene_name <- mapIds(
  org.Sc.sgd.db,
  keys = res_df$gene_id,
  column = "COMMON",
  keytype = "ORF",
  multiVals = "first"
)

# Add a description
# some genes will return NA
# many yeast open reading frames simply have no common name
# this is a fact about yeast rather than a code error
# keep the systematic identifier as the primary key
# common names change over time whereas systematic do not
res_df$description <- mapIds(
  org.Sc.sgd.db,
  keys = res_df$gene_id,
  column = "DESCRIPTION",
  keytype = "ORF",
  multiVals = "first"
)

# Checking top three significant IDs 
# ensure they mapped with common name and description as expected
head(res_df[order(res_df$padj), c("gene_id", "gene_name", "log2FoldChange","description")], 3)

##############################
### Functional enrichment ####
##############################

# GO enrichment compares our significant genes with a background list
# this is known as the gene universe
# This should include genes that had the opportunity to be identified
# (not filtered as NA in DESeq2)
# So the universe is every gene we tested, not every gene in the genome

# Compare the lengths of the gene universe versus sig_ids
# The universe should be thousands of genes and the significant list a few hundred
gene_universe <- res_df$gene_id[!is.na(res_df$padj)]
length(gene_universe)
length(sig_ids)

# The gene universe stays the same regardless of the thresholds used
# It is simply the genes we tested

### Run GO enrichment analysis

# ont = "BP" restricts the search to biological processes
# pAdjustMethod = "BH" is the same Benjamini-Hochberg correction
# ...because we are testing many hypotheses at once, one per GO term
# GeneRatio is the fraction of your significant genes in that process
# BgRatio is the fraction of the universe in that process
# A term is enriched when the first is much larger than the second

ego <- enrichGO(
  gene = sig_ids,
  universe = gene_universe,
  OrgDb = org.Sc.sgd.db,
  keyType = "ORF",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05
)

# Checking top of our GO enrichment output
head(as.data.frame(ego)[, c("Description", "GeneRatio", "BgRatio", "p.adjust")], 10)

# Visualise top ten terms
# A term with only three genes is rarely worth reporting
dotplot(ego, showCategory = 15) +
  labs(title = "Enriched biological processes: B vs A")

# Enrichment analysis described our gene list
# This gives us a starting point for exploring the biology
# but GO terms often overlap
# It does not show causation, and can only identify processes that have been annotated
# So we need to be careful not to overinterpret the results

### Running the analysis on up-regulated genes only

# derive up regulated, significant gene, filter for only tested genes, genes in universe
up_ids <- res_df$gene_id[res_df$padj < 0.05 & res_df$log2FoldChange > 1]
up_ids <- up_ids[!is.na(up_ids)]

# And now for downregulated genes, change the log2FoldChange to < -1
# Splitting by direction is almost always more informative than one combined list
down_ids <- res_df$gene_id[res_df$padj < 0.05 & res_df$log2FoldChange < -1]
down_ids <- down_ids[!is.na(down_ids)]

# non significant genes
neither_ids <- res_df$gene_id[!(res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1)]

# Run analysis for up regulated
ego_up <- enrichGO(
  gene = up_ids,
  universe = gene_universe,
  OrgDb = org.Sc.sgd.db,
  keyType = "ORF",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05)

# Run analysis for downregulated genes
ego_down <- enrichGO(
  gene = down_ids,
  universe = gene_universe,
  OrgDb = org.Sc.sgd.db,
  keyType = "ORF",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05)

# used only to test the enrichment processes is working correctly
neither_enrich <- enrichGO(
  gene = neither_ids,
  universe = gene_universe,
  OrgDb = org.Sc.sgd.db,
  keyType = "ORF",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05)

paste0("There were ", nrow(as.data.frame(ego_up)), " records returned by enrichGO for up regulated genes.")
paste0("There were ", nrow(as.data.frame(ego_down)), " records returned by enrichGO for down regulated genes.")  
paste0("There were ", nrow(as.data.frame(neither_enrich)), " records returned by enrichGO for non significant & low fold genes. This was performed as a sanity check for the zero up regulated gene enrichment.")

# Take a look
#head(as.data.frame(ego_up)[, c("Description", "GeneRatio", "BgRatio", "p.adjust")], 10)

# Visualise top ten terms for upregulated genes
#dotplot(ego_up, showCategory = 15) +
#  labs(title = "Enriched biological processes: B vs A in upregulated genes")

# Take a look
#head(as.data.frame(ego_down)[, c("Description", "GeneRatio", "BgRatio", "p.adjust")], 10)

# Visualise top ten terms for downregulated genes
dotplot(ego_down, showCategory = 15) +
  labs(title = "Enriched biological processes: B vs A in downregulated genes")
```

# 7. Your gene of interest

This section extracts and interprets the assigned gene of interest from the differential expression results. The gene's shrunken log₂ fold change, adjusted p-value, and significance classification are retrieved from the annotated DESeq2 results table. Gene Ontology annotations are then queried to identify biological processes associated with the gene, providing functional context for interpreting its expression changes between conditions A and B.

```{r goi}
params$gene_of_interest
# "YDL022W"

goi_row <- res_df[res_df$gene_id == params$gene_of_interest,]
goi_row

goi_lfc <- goi_row$log2FoldChange
goi_padj <- goi_row$padj
goi_status <- goi_row$status

goi_go <- AnnotationDbi::select(
  org.Sc.sgd.db,
  keys = params$gene_of_interest,
  keytype = "ORF",
  columns = c("GO", "ONTOLOGY")
)

goi_bp <- subset(goi_go, ONTOLOGY == "BP")
head(goi_bp)
```

# 8. Interpretation

**Gene of interest (max 4 sentences):**

The assigned gene of interest, YDL022W (GPD1), was significantly differentially expressed between the two conditions, with a shrunken log₂ fold change of -1.28, indicating lower expression in condition B relative to condition A. The result was highly significant (adjusted p-value = 1.79 × 10⁻¹³) and was therefore classified as Down in B. GPD1 encodes an NAD-dependent glycerol-3-phosphate dehydrogenase involved in glycerol biosynthesis and cellular responses to osmotic stress. These results indicate reduced expression of GPD1 in condition B and are consistent with differences in osmotic stress-related and glycerol biosynthetic processes between the two conditions.

**PCA (max 3 sentences):**

Principal component analysis (PCA) showed that samples clustered primarily according to experimental condition, indicating that condition A and condition B have distinct global gene expression profiles. Sample_026 showed additional variation along PC2 relative to the other A replicates but is grouped with condition A in PC1, indicating within-group variability rather than loss of condition separation. The clear separation between conditions supports the use of differential expression analysis to identify genes associated with the treatment effect.

**Enrichment (max 4 sentences):**

Gene Ontology over-representation analysis identified several enriched biological processes among genes that were down-regulated in condition B. The strongest enriched terms included oxoacid metabolic process, organic acid metabolic process, and cell wall organization or biogenesis, suggesting widespread changes to cellular metabolism and cell wall-related functions. Additional enriched terms were associated with purine nucleotide metabolism and carbohydrate transport, including hexose and fructose transmembrane transport. In contrast, no Biological Process terms were significantly enriched among genes up-regulated in condition B, indicating that the transcriptional response was dominated by coordinated repression of specific biological pathways rather than activation of a single strongly enriched process.

**Volcano caption (2 sentences):**

The most striking feature of the volcano plot is its asymmetry - more genes were down regulated in B (n=147) compared to up regulated in B (n=27). This means that relative to condition A, condition B is associated predominantly with *reduced* expression of many genes, while only a small number of genes show increased expression.

# 9. Session information

```{r session}
sessionInfo()
```