---
title: "SLE777 Assessment 4: Reproducible Differential Expression Report"
author: "Ivy Craw, Murray Pung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params: 
  group_id: "24"
  # Six assigned count files exactly as they appear in our allocation row
  condition_a_files: ["sample_024_counts.txt","sample_026_counts.txt","sample_029_counts.txt"]
  condition_b_files: ["sample_007_counts.txt","sample_017_counts.txt","sample_032_counts.txt"]
  counts_dir: "/home/shared/SLE777/A4_counts"
  gene_of_interest: "YDL022W"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 6)
set.seed(777)
# Load the packages you need. Record versions with sessionInfo() at the end.
library(clusterProfiler)
library(DESeq2)
library(ggplot2)
# What's magrittr used for?
library(magrittr)
library(apeglm)
library(pheatmap)
library(org.Sc.sgd.db)
library(AnnotationDbi)

# Suggested library green lit by unit chair for dynamic labelling
# OPTIONAL
library(ggrepel)

# Create figures directory if it does not already exist
dir.create("figures", showWarnings = FALSE)
```

This report analyses the cohort assigned to group ``r params$group_id``,
gene of interest ``r params$gene_of_interest``. 

# 1. Count matrix and metadata

```{r matrix}
####################################
# Build the count matrix
####################################

# Combine six HTSeq-count files into one gene six sample matrix
# Samples: condition A versus condition B, for assessment purposes

# To compare gene expression between conditions, we need all six samples side by side in one table
# The end product of this is a count matrix
# One row per gene and one column per sample and a read count in every cell
# We also build a sample metadata table
# These two objects are required for differential expression analysis (DESeq2)

### Descriptions of samples
# Counts provided for assessment purposes
# No additional information provided

####################################
# The sample metadata table
####################################

# Create sample metadata table with sample identifiers and condition
sample_metadata <- data.frame(
  sample_id = c("sample_024", "sample_026", "sample_029",
                "sample_007", "sample_017", "sample_032"),
  condition = c("A", "A", "A",
                "B", "B", "B")
)
sample_metadata

### Conditions should be factors
# Statistical models in R need categorical variables stored as factors
# At the moment condition is a character column
class(sample_metadata$condition)
str(sample_metadata)

# We want condition a to be the reference
# ...so that later a positive fold change will mean "higher in B than A"

# Convert the condition column to a factor with a as the reference level
sample_metadata$condition <- factor(
  sample_metadata$condition,
  levels = c("A", "B")
)
# Check that the conversion worked
levels(sample_metadata$condition)
# Check the structure of the metadata table
str(sample_metadata)

### Row names

# Name rows of the metadata after the samples
rownames(sample_metadata) <- sample_metadata$sample_id

# Confirm the metadata table is correct after these changes
sample_metadata

### Finding the count files

# Locate the count files assigned for the assessment
# We need to be specific here as there are many files
# Could also replace with params above to be safe
condition_A_files <- c(
  "sample_024_counts.txt",
  "sample_026_counts.txt",
  "sample_029_counts.txt"
)

condition_B_files <- c(
  "sample_007_counts.txt",
  "sample_017_counts.txt",
  "sample_032_counts.txt"
)

# Define the file path for our files
# This is the only time we are using an absolute path
# This was called for in the assessment instructions
#counts_dir <- "/home/shared/SLE777/A4_counts"
counts_dir <- "counts"

# Create vectors to save the files and their paths
count_files <- c(condition_A_files, condition_B_files)
count_paths <- file.path(counts_dir, count_files)

# Confirm that you can see the full file paths and file names
count_paths

# Safety checks to stop if not all six samples
stopifnot(length(count_files) == 6)
stopifnot(all(file.exists(count_paths)))

### Deriving sample names from file names

# Check file names again for what to strip
count_files

# Strip until we have sample identifiers only
sample_ids <- gsub("_counts\\.txt$", "", count_files)
sample_ids

# Check that every sample id derived from file name also appears in metadata table
# Stop the script if all results do not evaluate as TRUE
stopifnot(all(sample_ids %in% sample_metadata$sample_id))

####################################
# Merging the count files into one table
####################################

# Each count file is a two-column table: 
# ...a gene identifier, and the count for that one sample 
# To put six samples side by side
# ...we need to join these tables so that each row still refers to the same gene 

### Reading all six files with a loop
# The if else was added to the loop to create the counts_table object from scratch
# This deviates from the practical code as that included demonstration code
for (i in 1:length(count_paths)) {
  
  one_sample <- read.table(
    count_paths[i],
    header = FALSE,
    col.names = c("gene_id", sample_ids[i])
  )
  
  # Quick checks to reflect structure set earlier in code
  # Stop if the count table being processed does not have two columns
  # or if the first column is not character
  # or if the second column is not numeric
  stopifnot(
    ncol(one_sample) == 2,
    is.character(one_sample$gene_id),
    is.numeric(one_sample[[2]])
  )

  # Merge the new sample into the growing counts table
  # The first time through the loop the counts_table does not yet exist
  if (i == 1) {
    counts_table <- one_sample
  } else {
    counts_table <- merge(counts_table, one_sample, by = "gene_id")
  }
}

### Quick checks and guardrails for counts_table creation

# Confirm merged table contains one gene_id column plus six sample columns
stopifnot(ncol(counts_table) == length(sample_ids) + 1)

# Column names should be gene ID plus our sample IDs
expected_colnames <- c("gene_id", sample_ids)
stopifnot(all(colnames(counts_table) == expected_colnames))

# Check that the final table to see how many gene IDs and samples you have
# Remember that these are still inclusive of summary rows
message(
  "Merged count table created successfully: ",
  nrow(counts_table), " rows x ",
  ncol(counts_table), " columns."
)

####################################
# Separating the HTSeq-count summary rows
####################################

# Every count file has summary rows
# These were merged along with gene counts and will need to be separated

# Apply grepl to the wider table
is_summary <- grepl("^__", counts_table$gene_id)
sum(is_summary)

# Split table in two
htseq_summary <- counts_table[is_summary, ]
gene_table <- counts_table[!is_summary, ]

# Check that the split worked
dim(counts_table)

# Check that summary rows are in the summary table
dim(htseq_summary)

# Check that the gene table has no summary rows
dim(gene_table)

### Guardrails to be confident we have removed them

# Check that the summary table has five rows, one for each of the five HTSeq-count summary categories
stopifnot(nrow(htseq_summary) == 5)

# Check that the gene table has no summary rows
stopifnot(!any(grepl("^__", gene_table$gene_id)))

# And lastly, check that the gene table has the same number of rows as the original count files
# ...because we have not lost any genes in the merge
# The number of rows in the gene table should match the number of rows in any one of the original count files, minus the five summary rows
stopifnot(nrow(gene_table) == nrow(one_sample) - 5)

### These rows tell you how many reads were discarded because they fell outside any gene
# ...because they were ambiguous between overlapping genes
# or because their alignment quality was too low
# A sample with an unusually high __no_feature count deserves a closer look
htseq_summary

####################################
# Building the count matrix
####################################

# Need to remove gene_id so there is no text in table
# This is so that we can convert the table to a numeric matrix
count_matrix <- as.matrix(gene_table[, -1])
rownames(count_matrix) <- gene_table$gene_id

# Manually check the new object has the same dimensions minus one column
# Inspect visually the structure and first few rows
dim(count_matrix)
str(count_matrix)
head(count_matrix)

# And guardrailds for safety
stopifnot(
  nrow(count_matrix) == nrow(gene_table),
  ncol(count_matrix) == ncol(gene_table) - 1,
  all(rownames(count_matrix) == gene_table$gene_id),
  all(colnames(count_matrix) == sample_ids),
  is.matrix(count_matrix),
  is.numeric(count_matrix)
)

####################################
# Checking that the samples line up
####################################

# The columns of the count matrix and the rows of the metadata 
# ...must describe the same samples in the same order
# Nothing enforces this so we watch for silent errors

# First look at the two sets of names side by side
colnames(count_matrix)
sample_metadata$sample_id

# Now force the columns into the metadata order and check
count_matrix <- count_matrix[, sample_metadata$sample_id]
all(colnames(count_matrix) == sample_metadata$sample_id)
identical(colnames(count_matrix), rownames(sample_metadata))

# Also get R to check for us
stopifnot(all(colnames(count_matrix) == sample_metadata$sample_id))
stopifnot(identical(colnames(count_matrix), rownames(sample_metadata)))

### Always reorder the matrix columns to match the metadata rows
# ...always assert that they match
# ...and never continue past a failed check

####################################
# Library sizes and read assignments
####################################

# The library size of a sample is the total number of reads assigned to genes
# Samples with more reads will show higher counts for every gene
# ...which is why we cannot compare raw counts between samples
# Normalising converts counts into a quantity that does not depend on how deeply the sample was sequenced

# Remember to use colSums
# A lib size is a total per sample, and samples are columns in our matrix
library_sizes <- colSums(count_matrix)
length(library_sizes)
library_sizes

# Scale library sizes to millions of reads and round to 2 decimal places
round(library_sizes / 1e6, 2)

### Summary of reads that were not assigned to any gene

# Summary rows we set aside earlier
# Convert to matrix, removing header, then add up each column
summary_matrix <- as.matrix(htseq_summary[, -1])
rownames(summary_matrix) <- htseq_summary$gene_id
summary_matrix

# Sum the rows of the summary matrix to get the total number of reads that were not assigned to any gene
unassigned <- colSums(summary_matrix)
unassigned

### Calculate the proportion of reads assigned to genes

# First build a data frame
qc_table <- data.frame(
  sample_id = names(library_sizes),
  library_size = as.numeric(library_sizes),
  unassigned = as.numeric(unassigned[names(library_sizes)])
)
qc_table

# Add two more columns using existing columns
qc_table$total_reads <- qc_table$library_size + qc_table$unassigned
qc_table$percent_assigned <- 100 * qc_table$library_size / qc_table$total_reads

# Define sample order for plots so samples are grouped by condition
plot_sample_order <- sample_metadata$sample_id[
  order(sample_metadata$condition, sample_metadata$sample_id)
]

# Review table so far
qc_table

# A healthy RNA-seq sample from a well-annotated genome should assign a large majority of its reads to genes
# A sample that is markedly worse than the others is a candidate for exclusion
# Should a least be flagged in report

### Plotting the library sizes

# We need to merge condition alongside the library size for coloured bars
qc_table <- merge(qc_table, sample_metadata, by = "sample_id")

# Preserve the chosen sample order in ggplot
qc_table$sample_id <- factor(
  qc_table$sample_id,
  levels = plot_sample_order
)

# Review again
qc_table

# Fill aesthetics maps a column to bar colour
# theme_bw() function replaces the grey default background
# theme() call rotates the x-axis labels so that the long sample names do not overlap
ggplot(qc_table, aes(x = sample_id, y = library_size / 1e6, fill = condition)) +
  geom_col() +
  labs(
    title = "Library size per sample",
    x = "Sample",
    y = "Assigned reads (millions)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

### Plot percent_assigned for each sample as a bar chart, coloured by condition, with a sensible title and axis labels
ggplot(qc_table, aes(x = sample_id, y = percent_assigned, fill = condition)) +
  geom_col() +
  labs(
    title = "Percentage of reads assigned to genes",
    x = "Sample",
    y = "Reads assigned (%)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


####################################
# A first look across samples
####################################

# RNA-seq counts often have a skewed distribution
# Many genes have low counts while a smaller number have very high counts
# Log transformation makes the distributions easier to compare across samples

# Draw histogram of counts across samples
# Using the plot_sample_order var again to group samples
# The las = 2 argument turns the sample labels sideways so they fit
# Logarithm cannot take zero so we add 1 across whole matrix
boxplot(
  log10(count_matrix[, plot_sample_order] + 1),
  las = 2,
  ylab = "log10(count + 1)",
  main = "Distribution of gene counts per sample"
)

# Finally, check what working tables we've created thus far
ls()
```

# 2. Normalisation and PCA

To correct for sequencing depth, we divide each count by the sample's library size, before multiplying by one million. 

This is to make our counts more readable. 

Scaling like this, however, does not correct for gene length. For this reason, counts per million values cannot be compared between different genes. 

When we are looking to do differential expression, scaling our counts per million can be useful to compare the *same* gene across samples. 

```{r pca}
####################################
# Filtering genes with no expression
####################################

# Many genes are not expressed at all in this experiment
# They carry no information, they slow the analysis down
# ...and they cost us statistical power when we correct for multiple testing in Week 9

### Count how many genes have zero reads across every sample

# Create new var for total n of rows count as total number of genes
gene_totals <- rowSums(count_matrix)

### Remove the genes with zero reads across samples

# Create new variable to store filtered new matrix
# Keep all columns
count_matrix_filtered <- count_matrix[gene_totals > 0, ]

# Calculate summary statistics of our genes across samples
dim(count_matrix)

# Summary statistics for only genes with reads
dim(count_matrix_filtered)

##############################################
# Normalisation and exploring the experiment
##############################################

### Counts per million

# Each sample has its own library size
# Work through columns one at a time

# Copy count matrix so dimensions are the same
cpm <- count_matrix

# For each column i take col of counts
# divide it by that sample's lib size
# multiply by a million
# and put answer back in the count matrix copy cpm
for (i in 1:ncol(count_matrix)) {
  cpm[, i] <- count_matrix[, i] / library_sizes[i] * 1e6
}

# Check that we still have the same number genes and six samples
dim(cpm)
# Inspect results
head(cpm)
# and confirm that every column of CPM matrix sum is exactly one mill 1e+06
colSums(cpm)

### Filtering low-count genes

# We cannot reliably test genes with almost no reads anywhere
# They inflate the multiple-testing correction we apply later
# costing us the power to detect genes of interest

# A widely used rule is to keep a gene if it has a CPM above 1 in at least as many samples as the smallest group
# Our smallest group has three replicates
# table() will show us boolean TRUE FALSE values for how many genes passed this filter

keep <- rowSums(cpm > 1) >= 3
table(keep)

# Apply the filter to both matrices and retain all columns
# Check how many genes were dropped match the filter above
counts_filtered <- count_matrix[keep, ]
cpm_filtered <- cpm[keep, ]
dim(count_matrix)
dim(counts_filtered)

# Note that we keep the filtered raw counts as well
# DESeq2 next week wants raw counts, not CPM, because it models the counting process itself
# The CPM matrix is for exploration and plotting only

### Log transformation

# Count data are strongly skewed
# A handful of genes carry enormous counts and dominate every plot
# Taking logarithms compresses that range
# As always, we add 1 first, because the log of zero is undefined

log_cpm <- log2(cpm_filtered + 1)

# Inspect the log CPM matrix
dim(log_cpm)
head(log_cpm)

# Earlier in the srcipt we used log10 to squash the y-axis of a boxplot
# Here we switch to log2, and the change is deliberate
# Base 2 is conventional in genomics because a difference of 1 on the log2 scale means a doubling of expression
# ...a difference of 2 means a fourfold change, and so on
# That interpretation carries directly into the log2 fold changes we calculate later in the analysis
# For simply squashing a plot's axis either base works; for talking about fold changes, base 2 is the one that lets you read the number

### Did the normalisation work?

# Draw the distributions before and after to manually check that normalisation has worked
# Checking our samples are now comparable after normalisation
# Because both matrices are numeric, base R gives one box per sample from a single line of code

# Before normalisation
boxplot(
  log2(counts_filtered + 1),
  las = 2,
  ylab = "log2(count + 1)",
  main = "Before normalisation"
)

# After normalisation
boxplot(
  log_cpm,
  las = 2,
  ylab = "log2(CPM + 1)",
  main = "After normalisation"
)

##################################
# Principal component analysis ###
##################################

# Let's consider
# if the two stages really differ
# ...samples from the same stage should look more like each other than like samples from the other stage

# Our data have thousands of dimensions, one per gene, and we cannot draw that
# PCA finds a small number of new axes, principal components, that capture as much of the variation as possible
# It lets us show six samples measured on thousands of genes on an ordinary scatter plot
# PCA needs parts per million

# Each component is built from a combination of many genes
# PC1 is the direction along which the samples differ most
# PC2 the next most, and so on
# If the experiment worked, PC1 usually separates the conditions

### Transposing

# prcomp() expects observations x variables
# We have genes x samples, but our variables are genes
# So we need to switch! That's what t() does
# Run PCA to summarise the major patterns of variation between samples
pca <- prcomp(t(log_cpm))

# We should have the same number of PCA components as we do samples
dim(pca$x)
# Check that our row names have been switched to samples
rownames(pca$x)

# Add stop checks so R can check for us too
stopifnot(
  nrow(pca$x) == ncol(log_cpm),
  all(rownames(pca$x) == colnames(log_cpm))
)

# The sample coordinates live in pca$x
# ...so if I ran colnames(pca$x) I get "PC1" "PC2" "PC3" "PC4" "PC5" "PC6"

# Thousands of gene-expression measurements per sample → PCA 
# → a few summary dimensions (PC1, PC2, etc.) describing the major differences between samples

# PC1 → captures the most variation
# PC2 → captures the second most
# PC3 → captures the third most
# PC4 → captures the fourth most

# Look at how much percent variation each component captures
# prcomp() stores a standard deviation for each component in pca$sdev
# Variance is the square of standard deviation, so ^2
# Dividing each variance by the total gives each component's share
# Multiplying by 100 makes it a percentage
percent_var <- round(
  100 * pca$sdev^2 / sum(pca$sdev^2),
  1
)

# Inspect variation explained by the first four (highest-ranked) PCs
percent_var[1:4]

# Create a data frame containing each sample's coordinates
# on PC1 and PC2 (the two components explaining the most variation)
pca_df <- data.frame(
  sample_id = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2]
)

# Add sample metadata so PCA points can be identified by condition/stage
# sample_id is the shared key to identify PCA
pca_df <- merge(pca_df, sample_metadata, by = "sample_id")
pca_df

# ...note, a biologically important treatment effect can still be relatively small and therefore not obvious in those first two PCs

### Building the PCA figure with ggplot2
# The single most informative figure in a routine RNA-seq analysis

# Note, if the value comes from a column of your data, it goes inside aes()
# If you picked it yourself, it goes outside
# The same holds for colour, fill, size, shape and alpha

## Data frame
# We map to our PCA dataframe pca_df

## Aesthetic mapping
# Map our top PCA points to x and y, and condition to colour

## Geom
# We set the size of points with geom_point and draw
# We label points with sample_id using ggrepel rather than manually setting using vjust
# OPTIONAL ADDITION confirmed by unit chair to be acceptable

### scale_colour_manual
# Name them after the factor levels to the colour always follows the group
# Always use levels() e.g., pca_df$condition to check correct naming of data
pca_df$condition

## Labels
# Set ggplot object for PCA plot
# paste0() glues strings together without a separator

## Finally, theme_bw() just gives us a classic background colour scheme

# Draw the plot!
p_pca <- ggplot(
  pca_df,
  aes(x = PC1, y = PC2, colour = condition)
) +
  geom_point(size = 4) +
  geom_text_repel(
    aes(label = sample_id),
    size = 3,
    show.legend = FALSE
  ) +
  scale_colour_manual(
    values = c("A" = "#66c2a5", "B" = "#fc8d62")
  ) +
  labs(
    title = "PCA of normalised expression",
    x = paste0("PC1 (", percent_var[1], "% variance)"),
    y = paste0("PC2 (", percent_var[2], "% variance)"),
    colour = "Condition"
  ) +
  theme_bw()

# Draw object
p_pca

## IMPORTANT! 
# If the labels are confusing, try running that block again
# ggrepel will dynamically label the points so a reset will often fix funny positioning

# Save as png using ggsave()
# A resolution of 300 dpi is the usual minimum for a journal figure
ggsave(
  filename = "figures/pca_plot.png",
  plot = p_pca,
  width = 6,
  height = 5,
  dpi = 300
)

```

We can see that PC1 has separated our conditions and explains a large share of the variance, suggesting that the experiment has had a strong effect and differential expression will find plenty of genes. 

**Not sure if this wording is accurate here**

# 3. Differential expression

```{r deseq2}
#########################################
# Differential expression with DESeq2 ###
#########################################

# Week 9: Differential expression with DESeq2
# Input: filtered raw counts and sample metadata
# Output: a ranked table of differentially expressed genes.
# Comparison: b (45 days) vs a (10 days).

# if cols and rows have drifted apart we must stop and examine
# we set a as the first factor level earlier so it is the reference
stopifnot(all(colnames(counts_filtered) == sample_metadata$sample_id))

### 4. Why raw counts?
# A gene with 10 reads is proportionally far noisier than a gene with 10k reads
# One makes much more of a difference proportionately to ten than ten thousand
# DESeq2 insists on the raw integers and does its own normalisation internally

# Normalise for plotting
# Never normalise before a count-based statistical test
# DESeq2, edgeR and limma-voom all want raw counts and will normalise for you


### 5. Building the DESeqDataSet
# DESeq2 stores the counts, the metadata and the experimental design together in one object
# To ensure the three never drift apart

# 5.1 Build the object

# DESeqDataSetFromMatrix is a class, we pass it arguments
# countData is the gene by sample matrix of raw counts
# colData is the metadata, one row per column of the count matrix, in the same order
# design is a formula... ~ is modeled by, so model each gene's expression as a function of the condition column in the metadata
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered,
  colData = sample_metadata,
  design = ~ condition
)
dds

# 5.2 The design formula matters
# ie for batch experiments you would use design = ~ batch + condition
# The variable of interest goes last in the formula
# Everybg before it is a nuisance var to be adjusted for

# Check what DESeq2 bks the comparison is
# We set a as the first factor level, the reference, in w7
# A positive log fold change will mean higher in the b than the a
# R will set levels alphabetically if we don't explicitly tell it
levels(dds$condition)
resultsNames(dds)

############################
######### Step 1 ###########
############################

### 6. Size factors
# Dividing by the total library size is distorted by a handful of extremely highly expressed genes
# This is is a real risk in yeast where a few genes dominate the transcriptome

# DESeq2 is more clever and instead computes a size factor per sample
# a single number saying how deeply that sample was sequenced compared with a typical sample

dds <- estimateSizeFactors(dds)
sizeFactors(dds)

# A size factor near 1 means the sample is close to typical
# Above 1 means it was sequenced more deeply than average
# ... and DESeq2 will scale it down accordingly

# Comparing size factors with raw lib sizes from last week
# Should be correlated but not identical
library_sizes <- colSums(counts_filtered)
round(library_sizes / mean(library_sizes), 3)
round(sizeFactors(dds), 3)

# You can extract the normalised counts if you want to look at them
# Not used for testing
norm_counts <- counts(dds, normalized = TRUE)
head(round(norm_counts, 1))

############################
######### Step 2 ###########
############################

### 7. Dispersion
# Dispersion measures how much a gene's counts vary between replicates of the same condition
# ...beyond what the counting process alone would produce
# It is the parameter that decides whether an observed difference between groups is impressive or ordinary

# With only three replicates per group, estimating a gene's dispersion from its own data is hopeless
# DESeq2 solves this by looking at all genes together
# It fits a curve describing how dispersion typically depends on expression level
# ... then pulls each gene's noisy individual estimate towards that curve
# This is called shrinkage

dds <- estimateDispersions(dds)
plotDispEsts(dds)

# Black points are each gene's own estimate
# The red line is the fitted trend
# Blue points are the final, shrunken values actually used for testing

# Note, points circled in blue are genes with unusually high dispersion
# ...which DESeq2 leaves alone rather than shrinking
# because forcing them down would produce false positives

# The black cloud should scatter around the red line and blue should sit close
# A dispersion plot that does not look like this is a red flag

############################
####### Steps 3 and 4 ######
############################

### 8. Running the test
# The DESeq() function performs size factor estimation
# ...dispersion estimation and model fitting in one call

# We ran the first two steps separately so we could see them
# Running DESeq() now simply completes the job and won't redo work already done

dds <- DESeq(dds)

### DESeq should print our steps:
# using pre-existing size factors
# estimating dispersions
# found already estimated dispersions, replacing these
# gene-wise dispersion estimates
# mean-dispersion relationship
# final dispersion estimates
# fitting model and testing

# Always, always check levels before providing arguments
levels(dds$condition)

# Assign and check results
# The second name is the numerator and the third is the reference
# So it is critical we have b before a
res <- results(dds, contrast = c("condition", "B", "A"))
res

# The contrast reads as "condition: b versus a"
# A positive log fold change means higher in the b

############################
##### Reading results ######
############################

### 9. Reading the results table
# Look at the column descriptions that DESeq2 stores with the object

mcols(res)$description

# > mcols(res)$description
# "mean of normalized counts for all samples"      
# "log2 fold change (MLE): condition b vs a"
# "standard error: condition b vs a"        
# "Wald statistic: condition b vs a"        
# "Wald test p-value: condition b vs a"     
# "BH adjusted p-values"      

# Check the Week 9 content for a full breakdown

# Quick overview
summary(res)

### 10. Why the adjusted p-value?

# A p-value threshold of 0.05 means each gene has a 5 per cent chance of looking significant anyway

# Count how many genes clear the raw threshold
# sum(res$pvalue < 0.05, na.rm = TRUE)

# Now the adjusted threshold
# sum(res$padj < 0.05, na.rm = TRUE)

# The na.rm = TRUE argument is needed because DESeq2 sets padj to NA for genes it filtered out as untestable
# Without it, sum() would return NA

# pvalue <0.05
# If this gene were truly unchanged, there is only a 5 per cent chance I would see a difference this big
# Perfectly true for one gene, tested once
# Useless as a rule for choosing a gene list
# ...applying it thousands of times guarantees hundreds of mistakes

# padj < 0.05
# Of all the genes on my final list, about 5 per cent are expected to be false positives
# A statement about the list as a whole
# DESeq2 uses the Benjamini-Hochberg procedure to compute it, controlling the false discovery rate

# Always filter on padj 
# The raw p-value is shown so that you can inspect its distribution, which is a useful diagnostic
# but it is never the basis for a gene list

### 10.1 The p-value histogram
# A histogram of the raw p-values is one of the fastest diagnostics available

res_raw_df <- as.data.frame(res)

ggplot(res_raw_df, aes(x = pvalue)) +
  geom_histogram(binwidth = 0.02, boundary = 0) +
  labs(title = "Distribution of raw p-values", x = "p-value", y = "Number of genes") +
  theme_bw()

# We want a flat carpet aross most of the range
# With a spike near zero
# Flat bit is unchanged genes
# Spike is real signal
# A hist that rises towards 1 or that has bumps in the middle means the model is misspecified, do not trust

### 11. Shrinking the fold changes
# A gene with 4 reads in one group and 1 in the other has a log2 fold change of 2, which looks dramatic and means almost nobg
# Low-count genes produce wildly exaggerated fold changes

# Look at the problem first with an MA plot
# puts expression on the x-axis and fold change on the y-axis
# Every point is a gene

plotMA(res, ylim = c(-6, 6), main = "Before shrinkage")

# Notice the funnel shape on the left
# Weakly expressed genes fan out to extreme fold changes, because they are weakly expressed

# Shrinkage pulls those unreliable estimates towards zero while leaving well-measured genes almost untouched
# when an estimate is noisy, trust it less

# The coef argument needs the name DESeq2 gave the comparison
# Print the names and copy the one we want
resultsNames(dds)

# So we'll use condition_b_vs_a
res_shrunk <- lfcShrink(dds, coef = "condition_B_vs_A", type = "apeglm")
plotMA(res_shrunk, ylim = c(-6, 6), main = "After shrinkage")

# Use the shrunken fold changes for ranking genes, for plotting, and for deciding which genes are biologically interesting
# Use the unshrunken p-values and adjusted p-values for deciding which genes are statistically significant
# Shrinkage changes the effect sizes, not the significance

```

# 4. Significant genes

Here we are identifying the significant genes at padj < 0.05 and an absolute log2 fold change above 1, and splitting them by direction.

We rely on the following data objects returned by the DESeq2 analysis performed above:

- `dds` — not strictly needed for the basic filtering of significant genes, but useful downstream for plots/checks.
- `res` - the original, unshrunken DESeq2 B-vs-A results; useful for checking/testing and downstream analysis. 
- `res_shrunk` — shrunken results, which are used to construct our master data frame for identifying significant genes. 
- The correct coefficient used in lfcShrink() — it should come from resultsNames(dds) explicitly. 

```{r significant}
###################################
# Identifying significant genes ###
###################################

# Use the shrunken fold changes for ranking genes, for plotting, and for deciding which genes are biologically interesting
# Use the unshrunken p-values and adjusted p-values for deciding which genes are statistically significant
# Shrinkage changes the effect sizes, not the significance

# Turn the shrunken results into a plain data frame
# The third line reorders the columns so that gene_id comes first
res_df <- as.data.frame(res_shrunk)
res_df$gene_id <- rownames(res_shrunk)
res_df <- res_df[, c("gene_id", "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj")]
head(res_df)

# Remove the genes DESeq2 could not test, then sort by adjusted p-value
# !is.na() keeps the rows that have a real adjusted p-value
res_df <- res_df[!is.na(res_df$padj), ]
res_df <- res_df[order(res_df$padj), ]
head(res_df, 10)

# Apply both thresholds
# Statistical significance and a meaningful effect size are different requirements
# A good gene list demands both
# The abs() function takes the absolute value
# so this keeps genes that at least doubled in either direction
# Both conditions must hold
sig_genes <- res_df[res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1, ]

# Inspect the number of genes you get and the first few
nrow(sig_genes)
head(sig_genes)

# Split them by direction
# Remember that positive means higher in condition B, because A is the reference
up_in_b <- sig_genes[sig_genes$log2FoldChange > 0, ]
down_in_b <- sig_genes[sig_genes$log2FoldChange < 0, ]
nrow(up_in_b)
nrow(down_in_b)

# How many genes did DESeq2 leave untested
# so how many have padj set to NA?
sum(is.na(res$padj))

### Checking the result with our own eyes

# Take the top gene and plot its normalised counts in each sample
top_gene <- res_df$gene_id[1]
top_gene
plotCounts(dds, gene = top_gene, intgroup = "condition")

# The three A replicates should sit apart from the three B
# If they overlap, look at metadata closely to rule out labelling error

### Building the same figures in ggplot2 for a publishable version

### Calculate the top three genes

# This code is adapted from the unit content with suggestions from CoPilot
# Using index range from 1 to 3 top genes
# Because res_df is sorted by padj, indexing gives us the most significant genes
top_three_genes <- res_df$gene_id[1:3]
top_three_genes

# Extract normalised counts for the three most significant genes
# lapply(top_three_genes, function(gene) { runs function on each top gene
# gives us a list of three data frames so do.call(rbind, ...) stacks vertically
# it then runs the loop of plotcounts to get norm counts on each gene
top_three_counts <- do.call(
  rbind,
  lapply(top_three_genes, function(gene) {
    
    gene_norm <- plotCounts(
      dds,
      gene = gene,
      intgroup = "condition",
      returnData = TRUE
    )
    
    # Add sample IDs from the row names
    gene_norm$sample_id <- rownames(gene_norm)
    
    # Record which gene these counts belong to
    gene_norm$gene <- gene
    
    return(gene_norm)
  })
)

# Set gene factor levels to preserve significance order
top_three_counts$gene <- factor(
  top_three_counts$gene,
  levels = top_three_genes
)

# Plot all three genes
top_three_sig_genes <- ggplot(top_three_counts,
       aes(x = condition, y = count, colour = condition)) +
  geom_point(size = 4) +
  scale_y_log10() +
  scale_colour_manual(
    values = c("A" = "#66c2a5", "B" = "#fc8d62")
  ) +
  facet_wrap(~ gene) +
  labs(
    x = "Condition",
    y = "Normalised count"
  ) +
  theme_bw()

# OPTIONAL to include sample ids inside ggplot
#   geom_text_repel(aes(label = sample_id), size = 3, show.legend = FALSE) +
# But I think it's too squished once rendered

# Draw object
top_three_sig_genes

# OPTIONAL saving
ggsave(
  filename = "figures/top_three_sig_genes.png",
  plot = top_three_sig_genes,
  width = 6,
  height = 5,
  dpi = 300
)
```

# 5. Volcano plot and heatmap

```{r figures}

```

# 6. Annotation and GO enrichment

```{r enrichment}
#####################################
### 6. Annotating the gene names ####
#####################################

# Bioconductor supplies annotation packages that translate common name from systematic
# I.e., FMP16 vs YDR070C

# Check columns of library package
columns(org.Sc.sgd.db)

# Today we'll use common which returns the short name
# and description which returns a sentence describing what the gene does

# the mapIds() function looks up one identifier type and returns another
# keys are the identifiers we have
# keytype says what kind they are
# ORF for yeast systematic names
# column says what we want back
# multiVals = "first" handles the genes that map to more than one name by taking the first
res_df$gene_name <- mapIds(
  org.Sc.sgd.db,
  keys = res_df$gene_id,
  column = "COMMON",
  keytype = "ORF",
  multiVals = "first"
)
head(res_df[, c("gene_id", "gene_name", "log2FoldChange", "padj")], 10)

# Add a description
# some genes will return NA
# many yeast open reading frames simply have no common name
# this is a fact about yeast rather than a code error
# keep the systematic identifier as the primary key
# common names change over time whereas systematic do not
res_df$description <- mapIds(
  org.Sc.sgd.db,
  keys = res_df$gene_id,
  column = "DESCRIPTION",
  keytype = "ORF",
  multiVals = "first"
)
head(res_df$description, 3)

### Challenge 2
# print top 5 sig genes id, common name, and l2fc
# order rows by padj for significance, then choose the columns with a vector
head(res_df[order(res_df$padj), c("gene_id", "gene_name", "log2FoldChange")], 5)

#################################
### 7. Functional enrichment ####
#################################

# Gene Ontology over-representation analysis answers whether we have a random sample of the genome
# or if our listed genes are concentrated in particular biological processes
# For each bio process it asks whether more of our sig genes belong to that process
# ...than would be expected by chance

# Pathway enrichment analysis summarises the long gene list into a shorter
# ...interpret-able list of pathways
# Instead of having a list of a thousand genes you might get 50/60 bio pathways
# You can then check which genes are behind these pathways

# Three ingredients
# 1. Gene list of interest, ie diff expressed genes we want to summarise
# 2. List of background genes, eg all genes in human genome
# 3. List of gene sets, essentially groups of related genes
# ...eg if our list has a lot of genes related to breast cancer
# ...we'd need to tell our code which genes are involved in breast cancer

# We can then compare our list to the background list
# That gives us a list of overrepresented pathways
# e.g., our list of diff expressed genes involves a lot of genes involved in breast cancer

# say we know that ald involves inflammatory processes that involve pro-inflammatory cytokines, e.g., IL-6
# the question would be: is there an association between our genes diff expressed in alcoholic liver disease
# ...versus healthy cells, and IL-6 production
# in other words
# ...is our list of differentially expressed genes enriched with genes involved in IL-6 synthesis pathway

# We build a contingency table
# determine whether the fraction of genes of interest in the pathway is higher 
# ...compared to the fraction of genes outside the pathway, i.e., the background set

### Fishers exact test will give you a p-value for the pathway
# if the p-value is very low you can safely say that your list is overrepresented
# E.g., our gene list is enriched with genes involved in IL-6 production
# IL-6 production is an important pathway in alcoholic liver disease compared to healthy liver cells
# does not mean it's upregulated, we'd need to check gene list
# e.g., are those genes associated with the pos regulation of IL-6 production
# are they positive or negative regulators of that pathway

# multiple testing across pathways mean we will get strong results by chance, ie p value
# we correct this using multiple testing correction
# most commonly used method is Benjamini-Hochberg correction

# the p-value is the probability of seeing at least x n of genes out of n genes in our list
# annotated to a particular gene set term
# given the proportion of background genes that are annotated to that term

# the closer the p-value is to zero
# the more significant the particular gene set term associated with the group of genes is

# A common rule to select DEGs is p-val <0.005 and |FC| >2

### 7.1 The gene universe
# The comparison set is called the universe
# A gene that DESeq2 never tested (filtered out NA) could never have appeared on the list at all
# Counting such genes in the background makes the list look more special than it is
# So the universe is every gene we tested, not every gene in the genome

# Compare the lengths of the gene uni versus sig_ids
# The universe should be thousands of genes and the significant list a few hundred
gene_universe <- res_df$gene_id[!is.na(res_df$padj)]
length(gene_universe)
length(sig_ids)

# The principle does not change with the size of the effect
# State the universe you used and make it the genes you tested 
# A reviewer will ask

### 7.2 Running the analysis

# ont = "BP" restricts the search to biological processes
# pAdjustMethod = "BH" is the same Benjamini-Hochberg correction
# ...because we are testing many hypotheses at once, one per GO term
# GeneRatio is the fraction of your significant genes in that process
# BgRatio is the fraction of the universe in that process
# A term is enriched when the first is much larger than the second

ego <- enrichGO(
  gene = sig_ids,
  universe = gene_universe,
  OrgDb = org.Sc.sgd.db,
  keyType = "ORF",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05
)
head(as.data.frame(ego)[, c("Description", "GeneRatio", "BgRatio", "p.adjust")], 10)

# Visualise top ten terms
# The dot size shows how many genes are in each term
# Colour shows adjusted p-value
# A term with only three genes is rarely worth reporting
dotplot(ego, showCategory = 15) +
  labs(title = "Enriched biological processes: b vs a")

# Enrichment analysis described our gene list
# It shows that our gene list is concentrated in particular annotated processes
# ...more than a random list of the same size would be
# That is a useful summary and a good starting point for a hypothesis
# But its terms are highly redundant

# It does NOT show causation
# A process appearing in the list does not mean it drives the phenotype
# It cannot find a process nobody has annotated
# Poorly studied biology is invisible to this method, by construction

# Treat the output as a summary to be interpreted, not a result to be reported uncritically

### Challenge 3
# Running the analysis on up-regulated genes only

# Create our list
up_ids <- res_df$gene_id[res_df$padj < 0.05 & res_df$log2FoldChange > 1]
up_ids <- up_ids[!is.na(up_ids)]

# Run analysis
# Update the gene arg to point to up_ids
ego_up <- enrichGO(
  gene = up_ids,
  universe = gene_universe,
  OrgDb = org.Sc.sgd.db,
  keyType = "ORF",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05
)
head(as.data.frame(ego_up)[, c("Description", "GeneRatio", "BgRatio", "p.adjust")], 10)

# Visualise top ten terms for upregulated genes
dotplot(ego_up, showCategory = 15) +
  labs(title = "Enriched biological processes: B vs A in upregulated genes")

# And now for down regulated genes
# Create our list
# Change the log2FoldChange to < -1
# Splitting by direction is almost always more informative than one combined list
down_ids <- res_df$gene_id[res_df$padj < 0.05 & res_df$log2FoldChange < -1]
down_ids <- down_ids[!is.na(down_ids)]

# Run analysis
# Update the gene arg to point to down_ids
ego_down <- enrichGO(
  gene = down_ids,
  universe = gene_universe,
  OrgDb = org.Sc.sgd.db,
  keyType = "ORF",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05
)
head(as.data.frame(ego_down)[, c("Description", "GeneRatio", "BgRatio", "p.adjust")], 10)

# Visualise top ten terms for upregulated genes
dotplot(ego_down, showCategory = 15) +
  labs(title = "Enriched biological processes: B vs A in downregulated genes")

```

# 7. Your gene of interest

```{r goi}

```

# 8. Interpretation

**Gene of interest (max 4 sentences):**

**PCA (max 3 sentences):**

**Enrichment (max 4 sentences):**

**Volcano caption (2 sentences):**

# 9. Session information

```{r session}
sessionInfo()
```
