---
title: "SLE777 Assessment 4: Reproducible Differential Expression Report"
author: "Ivy Craw, Murray Pung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params: 
  group_id: "24"
  # Paste your six assigned count files exactly as they appear in your allocation row.
  condition_a_files: ["sample_080_counts.txt","sample_088_counts.txt","sample_096_counts.txt"]
  condition_b_files: ["sample_005_counts.txt","sample_036_counts.txt","sample_045_counts.txt"]
  counts_dir: "/home/shared/SLE777/A4_counts"
  gene_of_interest: "YBR126C"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 6)
set.seed(777)
# Load the packages you need. Record versions with sessionInfo() at the end.
library(clusterProfiler)
library(DESeq2)
library(ggplot2)
library(magrittr)
library(apeglm)
library(pheatmap)
library(org.Sc.sgd.db)
library(AnnotationDbi)

# Create figures directory if it does not already exist
dir.create("figures", showWarnings = FALSE)
```

This report analyses the cohort assigned to group ``r params$group_id``,
gene of interest ``r params$gene_of_interest``.

# 1. Count matrix and metadata

```{r matrix}
# Read six count files into one count matrix
condition_a_files <- c("sample_024_counts.txt", "sample_026_counts.txt", "sample_029_counts.txt")
condition_b_files <- c("sample_007_counts.txt", "sample_017_counts.txt", "sample_032_counts.txt")

counts_dir <- "../../shared/SLE777/A4_counts"

condition_a_files %in% dir(counts_dir)
condition_b_files %in% dir(counts_dir)

all_files <- c(condition_a_files, condition_b_files)
available_files <- intersect(all_files, dir(counts_dir))

# Build sample metadata
sample_metadata <- data.frame(
  sample_id = available_files,
  condition = ifelse(available_files %in% condition_a_files,"A","B"),stringsAsFactors = FALSE)
# Set condition A as reference level
sample_metadata$condition <- factor(
  sample_metadata$condition,
  levels = c("A", "B"))

rownames(sample_metadata) <- sample_metadata$sample_id

# Read count files and place in a list (which will be converted to the count matrix)
count_list <- lapply(
  available_files,
  function(f) {
    read.table(
      file.path(counts_dir, f),
      header = TRUE,
      sep = "\t",
      stringsAsFactors = FALSE)})

# Build count matrix
count_matrix <- data.frame(gene_id = count_list[[1]][, 1])

for (i in seq_along(count_list)) {
  count_matrix[[available_files[i]]] <- count_list[[i]][, 2]
}

rownames(count_matrix) <- count_matrix$gene_id
count_matrix$gene_id <- NULL

count_matrix <- as.matrix(count_matrix)
```

# 2. Normalisation and PCA

```{r pca}

```

# 3. Differential expression

```{r deseq2}

```

# 4. Significant genes

```{r significant}

```

# 5. Volcano plot and heatmap

```{r figures}

```

# 6. Annotation and GO enrichment

```{r enrichment}

```

# 7. Your gene of interest

```{r goi}

```

# 8. Interpretation

**Gene of interest (max 4 sentences):**

**PCA (max 3 sentences):**

**Enrichment (max 4 sentences):**

**Volcano caption (2 sentences):**

# 9. Session information

```{r session}
sessionInfo()
```
---
title: "SLE777 Assessment 4: Reproducible Differential Expression Report"
author: "Ivy Craw, Murray Pung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params: 
  group_id: "24"
  # Paste your six assigned count files exactly as they appear in your allocation row.
  condition_a_files: ["sample_080_counts.txt","sample_088_counts.txt","sample_096_counts.txt"]
  condition_b_files: ["sample_005_counts.txt","sample_036_counts.txt","sample_045_counts.txt"]
  counts_dir: "/home/shared/SLE777/A4_counts"
  gene_of_interest: "YBR126C"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 6)
set.seed(777)
# Load the packages you need. Record versions with sessionInfo() at the end.
library(ggplot2)
```

This report analyses the cohort assigned to group ``r params$group_id``,
gene of interest ``r params$gene_of_interest``.

# 1. Count matrix and metadata

```{r matrix}

```

# 2. Normalisation and PCA

```{r pca}

```

# 3. Differential expression

```{r deseq2}

```

# 4. Significant genes

```{r significant}

```

# 5. Volcano plot and heatmap

```{r figures}

```

# 6. Annotation and GO enrichment

```{r enrichment}

```

# 7. Your gene of interest

```{r goi}

```

# 8. Interpretation

**Gene of interest (max 4 sentences):**

**PCA (max 3 sentences):**

**Enrichment (max 4 sentences):**

**Volcano caption (2 sentences):**

# 9. Session information

```{r session}
sessionInfo()
```
