---
title: "SLE777 Assessment 4: Reproducible Differential Expression Report"
author: "Ivy Craw, Murray Pung"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params: 
  group_id: "24"
  # Paste your six assigned count files exactly as they appear in your allocation row.
  condition_a_files: ["sample_080_counts.txt","sample_088_counts.txt","sample_096_counts.txt"]
  condition_b_files: ["sample_005_counts.txt","sample_036_counts.txt","sample_045_counts.txt"]
  counts_dir: "/home/shared/SLE777/A4_counts"
  gene_of_interest: "YBR126C"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 6)
set.seed(777)
# Load the packages you need. Record versions with sessionInfo() at the end.
library(ggplot2)
```

This report analyses the cohort assigned to group ``r params$group_id``,
gene of interest ``r params$gene_of_interest``.

# 1. Count matrix and metadata

```{r matrix}

```

# 2. Normalisation and PCA

```{r pca}

```

# 3. Differential expression

```{r deseq2}

```

# 4. Significant genes

```{r significant}

```

# 5. Volcano plot and heatmap

```{r figures}

```

# 6. Annotation and GO enrichment

```{r enrichment}

```

# 7. Your gene of interest

```{r goi}

```

# 8. Interpretation

**Gene of interest (max 4 sentences):**

**PCA (max 3 sentences):**

**Enrichment (max 4 sentences):**

**Volcano caption (2 sentences):**

# 9. Session information

```{r session}
sessionInfo()
```
